---
title: "Urban/Rural and National Banking Lookbook"
author: "Codex"
date: 2026-09-01
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

# Purpose and definitions

This descriptive lookbook extends selected figures from the 2026 Small Business
Finance Report. It compares Wisconsin urban/metro and rural/nonmetro counties
and adds national benchmarks for the 50 states plus the District of Columbia.

“Urban” means USDA Rural-Urban Continuum Codes (RUCC) 1–3 (metropolitan), and
“rural” means RUCC 4–9 (nonmetropolitan). Current county codes use the 2023
classification; predecessor county codes absent from 2023 use the 2013 and then
2003 vintages. This is a fixed current-definition analysis rather than a claim
that a county's historical metropolitan status never changed.

```{r setup}
library(tidyverse)
library(readxl)
library(scales)
library(knitr)
library(glue)

project_root <- normalizePath(file.path("..", ".."), winslash = "/")
knitr::opts_knit$set(root.dir = project_root)

source(file.path(
  project_root,
  "1_code", "1_2_visualize", "figs_charlie", "_charlie_helpers.R"
))

required_file <- function(path) {
  if (!file.exists(path)) stop("Missing required input: ", path, call. = FALSE)
  path
}

num0 <- function(x) {
  replace_na(readr::parse_number(as.character(x)), 0)
}

safe_divide <- function(num, den) {
  ifelse(is.na(den) | den == 0, NA_real_, num / den)
}

short_dollar <- function(x) {
  case_when(
    abs(x) >= 1e9 ~ paste0("$", number(x / 1e9, accuracy = 0.1), "B"),
    abs(x) >= 1e6 ~ paste0("$", number(x / 1e6, accuracy = 0.01), "M"),
    TRUE ~ paste0("$", number(x / 1e3, accuracy = 1), "K")
  )
}

us_state_fips <- c(
  "01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13",
  "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25",
  "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36",
  "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48",
  "49", "50", "51", "53", "54", "55", "56"
)
us_state_abbr <- sort(c(state.abb, "DC"))

rurality_levels <- c("Urban/metro", "Rural/nonmetro")
rurality_colors <- c(
  "Urban/metro" = "#2E75B6",
  "Rural/nonmetro" = "#BF4D28"
)
comparison_levels <- c(
  "Wisconsin urban", "Wisconsin rural", "U.S. urban", "U.S. rural"
)
comparison_colors <- c(
  "Wisconsin urban" = "#2E75B6",
  "Wisconsin rural" = "#BF4D28",
  "U.S. urban" = "#4C956C",
  "U.S. rural" = "#7B5EA7"
)
bucket_levels <- c("Under $100K", "$100K–$250K", "$250K–$1M")

stopifnot(
  length(us_state_fips) == 51L,
  n_distinct(us_state_fips) == 51L,
  length(us_state_abbr) == 51L,
  "55" %in% us_state_fips,
  "WI" %in% us_state_abbr
)
```

# Classification and population audits

## RUCC precedence lookup

```{r rucc-lookup}
rucc_2023 <- read_excel(required_file(
  file.path("0_inputs", "Ruralurbancontinuumcodes2023.xlsx")
)) |>
  transmute(
    county_fips = str_pad(as.character(FIPS), 5, pad = "0"),
    rucc_2023 = as.integer(RUCC_2023)
  ) |>
  filter(str_sub(county_fips, 1, 2) %in% us_state_fips)

rucc_2013 <- read_excel(required_file(
  file.path("0_inputs", "Ruralurbancontinuumcodes2013.xls")
)) |>
  transmute(
    county_fips = str_pad(as.character(FIPS), 5, pad = "0"),
    rucc_2013 = as.integer(RUCC_2013)
  ) |>
  filter(str_sub(county_fips, 1, 2) %in% us_state_fips)

rucc_2003 <- read_excel(required_file(
  file.path("0_inputs", "Ruralurbancontinuumcodes2003.xls")
)) |>
  transmute(
    county_fips = str_pad(as.character(`FIPS Code`), 5, pad = "0"),
    rucc_2003 = as.integer(`2003 Rural-urban Continuum Code`)
  ) |>
  filter(str_sub(county_fips, 1, 2) %in% us_state_fips)

stopifnot(
  !anyDuplicated(rucc_2023$county_fips),
  !anyDuplicated(rucc_2013$county_fips),
  !anyDuplicated(rucc_2003$county_fips),
  all(rucc_2023$rucc_2023 %in% 1:9),
  all(rucc_2013$rucc_2013 %in% 1:9),
  all(rucc_2003$rucc_2003 %in% 1:9)
)

rucc_lookup <- full_join(rucc_2023, rucc_2013, by = "county_fips") |>
  full_join(rucc_2003, by = "county_fips") |>
  mutate(
    rucc = coalesce(rucc_2023, rucc_2013, rucc_2003),
    rucc_source_vintage = case_when(
      !is.na(rucc_2023) ~ "2023",
      !is.na(rucc_2013) ~ "2013 fallback",
      !is.na(rucc_2003) ~ "2003 fallback",
      TRUE ~ NA_character_
    ),
    rurality = factor(
      if_else(rucc <= 3L, "Urban/metro", "Rural/nonmetro"),
      levels = rurality_levels
    )
  ) |>
  select(county_fips, rucc, rucc_source_vintage, rurality)

stopifnot(
  !anyDuplicated(rucc_lookup$county_fips),
  !anyNA(rucc_lookup$rucc),
  all(rucc_lookup$rucc %in% 1:9)
)

rucc_audit <- rucc_lookup |>
  count(rucc_source_vintage, name = "county_keys") |>
  arrange(factor(rucc_source_vintage,
                 levels = c("2023", "2013 fallback", "2003 fallback")))

kable(rucc_audit, caption = "County keys by RUCC source vintage")
```

South Boston independent city (`51780`) appears in the 2000–2002 CRA files
despite its 1995 merger into Halifax County. Those numerator records are mapped
to Halifax County (`51083`) before classification.

## Annual subgroup populations

The denominator is resident population living in the same RUCC subgroup as the
loan numerator. It uses final intercensal county estimates for 2000–2019 and
Census Vintage 2025 county estimates for 2020–2023.

```{r population-panel}
pep_dir <- file.path(
  "0_inputs", "data_charlie", "Census", "PEP_county_population"
)

pep_2000_raw <- read_csv(
  required_file(file.path(pep_dir, "co-est00int-tot.csv")),
  show_col_types = FALSE,
  progress = FALSE
)

pep_2000 <- pep_2000_raw |>
  filter(SUMLEV == 50, str_pad(as.character(STATE), 2, pad = "0") %in% us_state_fips) |>
  transmute(
    state_fips = str_pad(as.character(STATE), 2, pad = "0"),
    county_fips = str_c(
      state_fips,
      str_pad(as.character(COUNTY), 3, pad = "0")
    ),
    across(POPESTIMATE2000:POPESTIMATE2009)
  ) |>
  pivot_longer(
    starts_with("POPESTIMATE"),
    names_pattern = "POPESTIMATE(\\d{4})",
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(year = as.integer(year), source_period = "2000–2010 intercensal")

pep_2010 <- read_csv(
  required_file(file.path(pep_dir, "cc-est2020int-agesex-all.csv")),
  show_col_types = FALSE,
  progress = FALSE
) |>
  filter(
    SUMLEV == 50,
    YEAR %in% 2:11,
    str_pad(as.character(STATE), 2, pad = "0") %in% us_state_fips
  ) |>
  transmute(
    year = as.integer(YEAR) + 2008L,
    state_fips = str_pad(as.character(STATE), 2, pad = "0"),
    county_fips = str_c(
      state_fips,
      str_pad(as.character(COUNTY), 3, pad = "0")
    ),
    population = as.numeric(POPESTIMATE),
    source_period = "2010–2020 intercensal"
  )

pep_2020_raw <- read_csv(
  required_file(file.path(pep_dir, "co-est2025-alldata.csv")),
  col_types = cols(
    SUMLEV = col_character(), STATE = col_character(), COUNTY = col_character(),
    .default = col_guess()
  ),
  progress = FALSE
)

pep_2020 <- pep_2020_raw |>
  filter(SUMLEV == "050", STATE %in% us_state_fips) |>
  transmute(
    state_fips = STATE,
    county_fips = str_c(STATE, COUNTY),
    across(POPESTIMATE2020:POPESTIMATE2023)
  ) |>
  pivot_longer(
    starts_with("POPESTIMATE"),
    names_pattern = "POPESTIMATE(\\d{4})",
    names_to = "year",
    values_to = "population"
  ) |>
  mutate(year = as.integer(year), source_period = "2025 vintage")

county_population_unclassified <- bind_rows(pep_2000, pep_2010, pep_2020) |>
  arrange(year, county_fips)

population_unmatched <- county_population_unclassified |>
  distinct(county_fips) |>
  anti_join(rucc_lookup, by = "county_fips")

county_population <- county_population_unclassified |>
  left_join(rucc_lookup, by = "county_fips")

stopifnot(
  nrow(population_unmatched) == 0L,
  !anyDuplicated(county_population[c("year", "county_fips")]),
  identical(sort(unique(county_population$year)), 2000:2023),
  all(county_population$population > 0),
  !anyNA(county_population$rurality)
)

state_subgroup_population <- county_population |>
  group_by(year, state_fips, rurality) |>
  summarise(population = sum(population), .groups = "drop")

population_series <- bind_rows(
  state_subgroup_population |>
    filter(state_fips == "55") |>
    transmute(year, geography = "Wisconsin", rurality, population),
  state_subgroup_population |>
    group_by(year, rurality) |>
    summarise(population = sum(population), .groups = "drop") |>
    mutate(geography = "United States", .before = rurality)
) |>
  mutate(
    series = factor(
      case_when(
        geography == "Wisconsin" & rurality == "Urban/metro" ~ "Wisconsin urban",
        geography == "Wisconsin" ~ "Wisconsin rural",
        geography == "United States" & rurality == "Urban/metro" ~ "U.S. urban",
        TRUE ~ "U.S. rural"
      ),
      levels = comparison_levels
    )
  )

population_reconciliation <- county_population |>
  group_by(year) |>
  summarise(
    counties_or_equivalents = n(),
    input_population = sum(population),
    classified_population = sum(population[!is.na(rurality)]),
    difference = classified_population - input_population,
    .groups = "drop"
  )

stopifnot(
  nrow(population_series) == 24L * 4L,
  all(population_series$population > 0),
  all(population_reconciliation$difference == 0)
)

kable(
  population_reconciliation |>
    filter(year %in% c(2000L, 2009L, 2010L, 2019L, 2020L, 2023L)) |>
    mutate(across(ends_with("population"), comma)),
  caption = "County-population classification reconciliation"
)
```

# Harmonized CRA county panel

CRA source formats change twice between 2000 and 2023. The loader retains
county FIPS through classification and converts amounts reported in thousands
of dollars to dollars in 2000–2004 and 2019–2023.

```{r cra-county-loader}
load_cra_county_year <- function(year) {
  if (year <= 2004L) {
    raw <- read_delim(
      required_file(file.path(
        charlie_cra_dir, "aggr", sprintf("tract_%d.txt", year)
      )),
      delim = "|",
      show_col_types = FALSE,
      progress = FALSE
    ) |>
      filter(report_level == 200) |>
      transmute(
        state_fips = str_pad(as.character(state), 2, pad = "0"),
        county_fips_original = str_c(
          state_fips,
          str_pad(as.character(county), 3, pad = "0")
        ),
        loans_u100k_num = num0(num_100k),
        loans_100_250_num = num0(num_250k),
        loans_250_1m_num = num0(num_1mil),
        loans_u100k_amt = num0(vol_100k) * 1000,
        loans_100_250_amt = num0(vol_250k) * 1000,
        loans_250_1m_amt = num0(vol_1mil) * 1000
      )
  } else if (year <= 2018L) {
    raw <- read_csv(
      required_file(file.path(
        charlie_cra_dir, "cra_old", sprintf("cra_%d.csv", year)
      )),
      show_col_types = FALSE,
      progress = FALSE
    ) |>
      transmute(
        county_fips_original = str_pad(as.character(fips), 5, pad = "0"),
        state_fips = str_sub(county_fips_original, 1, 2),
        loans_u100k_num = num0(loan_count_100k),
        loans_100_250_num = num0(loan_count_250k),
        loans_250_1m_num = num0(loan_count_1M),
        loans_u100k_amt = num0(loan_vol_100k),
        loans_100_250_amt = num0(loan_vol_250k),
        loans_250_1m_amt = num0(loan_vol_1M)
      )
  } else {
    raw <- read_fwf(
      required_file(file.path(
        charlie_cra_dir,
        sprintf("%dexp_aggr", year - 2000L),
        sprintf("cra%d_Aggr_A11.dat", year)
      )),
      col_positions = fwf_widths(cra_fixed_widths, cra_fixed_names),
      show_col_types = FALSE,
      progress = FALSE
    ) |>
      filter(as.character(Report_Level) == "200") |>
      transmute(
        state_fips = str_pad(as.character(State_FIPS), 2, pad = "0"),
        county_fips_original = str_c(
          state_fips,
          str_pad(as.character(County_FIPS), 3, pad = "0")
        ),
        loans_u100k_num = num0(Loans_U100k_Num),
        loans_100_250_num = num0(Loans_100_250_Num),
        loans_250_1m_num = num0(Loans_250_1M_Num),
        loans_u100k_amt = num0(Loans_U100k_Amt) * 1000,
        loans_100_250_amt = num0(Loans_100_250_Amt) * 1000,
        loans_250_1m_amt = num0(Loans_250_1M_Amt) * 1000
      )
  }

  raw |>
    filter(state_fips %in% us_state_fips) |>
    mutate(
      year = as.integer(year),
      county_fips = if_else(
        county_fips_original == "51780",
        "51083",
        county_fips_original
      )
    ) |>
    group_by(year, state_fips, county_fips_original, county_fips) |>
    summarise(across(starts_with("loans_"), sum), .groups = "drop")
}

cra_county_unclassified <- map_dfr(2000:2023, load_cra_county_year)

cra_unmatched <- cra_county_unclassified |>
  distinct(county_fips) |>
  anti_join(rucc_lookup, by = "county_fips")

cra_county <- cra_county_unclassified |>
  left_join(rucc_lookup, by = "county_fips")

cra_value_cols <- names(cra_county)[str_detect(names(cra_county), "^loans_")]

stopifnot(
  nrow(cra_unmatched) == 0L,
  identical(sort(unique(cra_county$year)), 2000:2023),
  !anyNA(cra_county$rurality),
  all(as.matrix(cra_county[cra_value_cols]) >= 0),
  all(cra_county$state_fips %in% us_state_fips)
)

cra_subgroups <- bind_rows(
  cra_county |>
    filter(state_fips == "55") |>
    group_by(year, rurality) |>
    summarise(across(all_of(cra_value_cols), sum), .groups = "drop") |>
    mutate(geography = "Wisconsin", .before = rurality),
  cra_county |>
    group_by(year, rurality) |>
    summarise(across(all_of(cra_value_cols), sum), .groups = "drop") |>
    mutate(geography = "United States", .before = rurality)
) |>
  mutate(
    series = factor(
      case_when(
        geography == "Wisconsin" & rurality == "Urban/metro" ~ "Wisconsin urban",
        geography == "Wisconsin" ~ "Wisconsin rural",
        geography == "United States" & rurality == "Urban/metro" ~ "U.S. urban",
        TRUE ~ "U.S. rural"
      ),
      levels = comparison_levels
    )
  )

cra_total_check <- cra_county |>
  group_by(year, geography = if_else(state_fips == "55", "Wisconsin", "United States")) |>
  summarise(across(all_of(cra_value_cols), sum), .groups = "drop")

wi_frequency_reference <- load_cra_wi_frequency_series()
wi_volume_reference <- load_cra_wi_volume_series()

wi_generic <- cra_county |>
  filter(state_fips == "55") |>
  group_by(year) |>
  summarise(across(all_of(cra_value_cols), sum), .groups = "drop")

frequency_match <- isTRUE(all.equal(
  wi_generic |>
    select(year, loans_u100k_num, loans_100_250_num, loans_250_1m_num),
  wi_frequency_reference |>
    select(year, loans_u100k_num, loans_100_250_num, loans_250_1m_num),
  check.attributes = FALSE
))

volume_match <- isTRUE(all.equal(
  wi_generic |>
    select(year, loans_u100k_amt, loans_100_250_amt, loans_250_1m_amt),
  wi_volume_reference |>
    select(year, loans_u100k_amt, loans_100_250_amt, loans_250_1m_amt),
  check.attributes = FALSE
))

cra_loader_audit <- tibble(
  check = c(
    "All county-year records have RUCC",
    "Wisconsin loan counts reproduce Figure 8 loader",
    "Wisconsin loan volumes reproduce Figure 9 loader",
    "Legacy South Boston records successor-mapped"
  ),
  passed = c(
    nrow(cra_unmatched) == 0L,
    frequency_match,
    volume_match,
    sum(cra_county_unclassified$county_fips_original == "51780") == 3L
  )
)

stopifnot(all(cra_loader_audit$passed))
kable(cra_loader_audit, caption = "CRA loader and geography audits")
```

## Subgroup normalization

```{r cra-indexes}
cra_normalized <- cra_subgroups |>
  left_join(
    population_series |>
      select(year, geography, rurality, population),
    by = c("year", "geography", "rurality")
  ) |>
  pivot_longer(
    cols = all_of(cra_value_cols),
    names_to = "measure_bucket",
    values_to = "raw_value"
  ) |>
  mutate(
    measure = if_else(str_ends(measure_bucket, "_num"), "Frequency", "Volume"),
    bucket = case_when(
      str_detect(measure_bucket, "u100k") ~ "Under $100K",
      str_detect(measure_bucket, "100_250") ~ "$100K–$250K",
      TRUE ~ "$250K–$1M"
    ),
    bucket = factor(bucket, levels = bucket_levels),
    per_10k = safe_divide(raw_value, population) * 10000
  ) |>
  group_by(series, measure, bucket) |>
  arrange(year, .by_group = TRUE) |>
  mutate(index = per_10k / first(per_10k) * 100) |>
  ungroup()

index_audit <- cra_normalized |>
  group_by(series, measure, bucket) |>
  summarise(
    years = n(),
    first_year = min(year),
    last_year = max(year),
    base_index = first(index),
    positive_denominators = all(population > 0),
    .groups = "drop"
  )

stopifnot(
  nrow(index_audit) == 4L * 2L * 3L,
  all(index_audit$years == 24L),
  all(index_audit$first_year == 2000L),
  all(index_audit$last_year == 2023L),
  all(abs(index_audit$base_index - 100) < 1e-10),
  all(index_audit$positive_denominators)
)
```

# Urban/rural heterogeneity

## Wisconsin branches by county rurality

```{r fdic-data}
fdic_yearly <- map_dfr(2000:2024, function(year) {
  sod <- load_fdic_sod(year) |>
    mutate(
      county_fips = str_pad(as.character(STCNTYBR), 5, pad = "0"),
      branch_state = as.character(STALPBR)
    )

  wi_classified <- sod |>
    filter(branch_state == "WI") |>
    left_join(rucc_lookup, by = "county_fips")

  if (anyNA(wi_classified$rurality)) {
    stop("Unclassified Wisconsin FDIC county FIPS in ", year, call. = FALSE)
  }

  wi_counts <- wi_classified |>
    count(rurality, name = "branches") |>
    complete(
      rurality = factor(rurality_levels, levels = rurality_levels),
      fill = list(branches = 0L)
    ) |>
    mutate(year = year, .before = rurality)

  national <- sod |>
    filter(branch_state %in% us_state_abbr)

  bind_rows(
    wi_counts |>
      mutate(metric = "wi_rurality", institutions = NA_integer_),
    tibble(
      year = year,
      rurality = factor(NA_character_, levels = rurality_levels),
      branches = nrow(national),
      metric = "national_total",
      institutions = n_distinct(national$CERT)
    )
  )
})

wi_branch_rurality <- fdic_yearly |>
  filter(metric == "wi_rurality") |>
  select(year, rurality, branches)

wi_branch_share <- wi_branch_rurality |>
  group_by(year) |>
  mutate(total_branches = sum(branches)) |>
  ungroup() |>
  filter(rurality == "Rural/nonmetro") |>
  transmute(year, rural_share = branches / total_branches)

national_fdic <- fdic_yearly |>
  filter(metric == "national_total") |>
  select(year, institutions, branches)

wi_fdic_total <- map_dfr(c(2000L, 2024L), function(year) {
  wi <- load_fdic_sod(year) |> filter(STALPBR == "WI")
  tibble(year, institutions = n_distinct(wi$CERT), branches = nrow(wi))
})

stopifnot(
  nrow(wi_branch_rurality) == 25L * 2L,
  all(wi_branch_share$rural_share >= 0 & wi_branch_share$rural_share <= 1),
  identical(wi_fdic_total$institutions, c(365L, 193L)),
  identical(wi_fdic_total$branches, c(2116L, 1661L)),
  all(national_fdic$branches > 0),
  all(national_fdic$institutions > 0)
)
```

```{r wi-branch-rurality-figure, fig.width=10, fig.height=5}
ggplot(
  wi_branch_rurality,
  aes(x = year, y = branches, color = rurality, shape = rurality)
) +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.1) +
  scale_color_manual(values = rurality_colors) +
  scale_shape_manual(values = c("Urban/metro" = 16, "Rural/nonmetro" = 15)) +
  scale_x_continuous(breaks = seq(2000, 2024, 2)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "Wisconsin Bank Branches by County Rurality",
    subtitle = "Branch location classified with RUCC 1–3 versus 4–9",
    x = "Year", y = "Branches", color = NULL, shape = NULL,
    caption = "Data: FDIC Summary of Deposits; USDA Rural-Urban Continuum Codes"
  ) +
  charlie_theme() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

```{r wi-rural-branch-share-figure, fig.width=10, fig.height=4.8}
ggplot(wi_branch_share, aes(x = year, y = rural_share)) +
  geom_line(color = rurality_colors[["Rural/nonmetro"]], linewidth = 1.1) +
  geom_point(color = rurality_colors[["Rural/nonmetro"]], size = 2.2) +
  scale_x_continuous(breaks = seq(2000, 2024, 2)) +
  scale_y_continuous(labels = label_percent(accuracy = 1), limits = c(0, NA)) +
  labs(
    title = "Rural Share of Wisconsin Bank Branches",
    subtitle = "Share of classified Wisconsin branch locations in RUCC 4–9 counties",
    x = "Year", y = "Rural share",
    caption = "Data: FDIC Summary of Deposits; USDA Rural-Urban Continuum Codes"
  ) +
  charlie_theme() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r wi-branch-endpoints}
kable(
  wi_branch_rurality |>
    filter(year %in% c(2000L, 2024L)) |>
    left_join(wi_branch_share, by = "year") |>
    mutate(rural_share = percent(rural_share, accuracy = 0.1)),
  caption = "Wisconsin branch rurality audit (endpoint years)"
)
```

## Rural Wisconsin lending by original loan size, 2023

```{r figure7-helper}
build_figure7_table <- function(df) {
  out <- tibble(
    Category = factor(bucket_levels, levels = bucket_levels),
    loans = c(
      sum(df$loans_u100k_num),
      sum(df$loans_100_250_num),
      sum(df$loans_250_1m_num)
    ),
    volume = c(
      sum(df$loans_u100k_amt),
      sum(df$loans_100_250_amt),
      sum(df$loans_250_1m_amt)
    )
  ) |>
    mutate(
      loan_share = loans / sum(loans),
      volume_share = volume / sum(volume),
      average_size = safe_divide(volume, loans)
    )

  bind_rows(
    out,
    tibble(
      Category = factor("Total", levels = c(bucket_levels, "Total")),
      loans = sum(out$loans),
      volume = sum(out$volume),
      loan_share = sum(out$loan_share),
      volume_share = sum(out$volume_share),
      average_size = safe_divide(sum(out$volume), sum(out$loans))
    )
  ) |>
    mutate(Category = factor(as.character(Category), levels = c(bucket_levels, "Total")))
}

format_figure7 <- function(df) {
  df |>
    transmute(
      Category,
      `# of Loans` = comma(loans, accuracy = 1),
      `% of Loans` = percent(loan_share, accuracy = 0.1),
      Volume = dollar(volume, accuracy = 1, scale_cut = cut_short_scale()),
      `% of Volume` = percent(volume_share, accuracy = 0.1),
      `Avg. Size` = dollar(average_size, accuracy = 1)
    )
}

rural_wi_2023 <- cra_county |>
  filter(
    year == 2023L,
    state_fips == "55",
    rurality == "Rural/nonmetro"
  )

rural_wi_figure7 <- build_figure7_table(rural_wi_2023)

stopifnot(
  nrow(rural_wi_figure7) == 4L,
  abs(sum(rural_wi_figure7$loan_share[1:3]) - 1) < 1e-10,
  abs(sum(rural_wi_figure7$volume_share[1:3]) - 1) < 1e-10,
  all(rural_wi_2023$rucc %in% 4:9)
)

kable(
  format_figure7(rural_wi_figure7),
  align = c("l", "r", "r", "r", "r", "r"),
  caption = "Rural Wisconsin CRA small-business lending by original loan size (2023)"
)
```

## Wisconsin credit-union commercial loan size by headquarters rurality

Credit unions are classified by headquarters county. The plotted average is a
ratio of group totals: total commercial-loan amount divided by total
commercial-loan count.

```{r ncua-data}
normalize_county_name <- function(x) {
  x |>
    str_to_upper() |>
    str_replace_all("\\.", "") |>
    str_replace(" COUNTY$", "") |>
    str_squish()
}

wi_rucc_name_lookup <- read_excel(required_file(
  file.path("0_inputs", "Ruralurbancontinuumcodes2023.xlsx")
)) |>
  filter(State == "WI") |>
  transmute(
    county_key = normalize_county_name(County_Name),
    county_fips = str_pad(as.character(FIPS), 5, pad = "0"),
    rucc = as.integer(RUCC_2023),
    rurality = factor(
      if_else(rucc <= 3L, "Urban/metro", "Rural/nonmetro"),
      levels = rurality_levels
    )
  )

ncua_results <- map(2017:2024, function(year) {
  main_offices <- load_ncua_branch(year) |>
    filter(MainOffice == "Yes") |>
    distinct(CU_NUMBER, .keep_all = TRUE)

  wi_main <- main_offices |>
    filter(PhysicalAddressStateCode == "WI") |>
    mutate(county_key = normalize_county_name(PhysicalAddressCountyName)) |>
    left_join(wi_rucc_name_lookup, by = "county_key")

  if (anyNA(wi_main$rurality)) {
    missing_names <- wi_main |>
      filter(is.na(rurality)) |>
      distinct(PhysicalAddressCountyName) |>
      pull()
    stop(
      "Unmatched Wisconsin NCUA headquarters county names in ", year, ": ",
      paste(missing_names, collapse = ", "),
      call. = FALSE
    )
  }

  fs <- load_ncua_fs(year) |>
    transmute(
      CU_NUMBER,
      commercial_amount = num0(ACCT_475A1),
      commercial_count = num0(ACCT_090A1)
    )

  if (anyDuplicated(fs$CU_NUMBER)) {
    stop("Duplicate NCUA financial-statement CU_NUMBER in ", year, call. = FALSE)
  }

  wi_groups <- wi_main |>
    select(CU_NUMBER, rurality) |>
    inner_join(fs, by = "CU_NUMBER") |>
    group_by(rurality) |>
    summarise(
      amount = sum(commercial_amount),
      count = sum(commercial_count),
      credit_unions = n_distinct(CU_NUMBER),
      .groups = "drop"
    ) |>
    mutate(year = year, average_size = safe_divide(amount, count), .before = rurality)

  eligible_main <- main_offices |>
    filter(PhysicalAddressStateCode %in% us_state_abbr)

  geo_groups <- bind_rows(
    wi_main |>
      transmute(CU_NUMBER, geography = "Wisconsin"),
    eligible_main |>
      transmute(CU_NUMBER, geography = "United States")
  ) |>
    inner_join(fs, by = "CU_NUMBER") |>
    group_by(geography) |>
    summarise(
      amount = sum(commercial_amount),
      count = sum(commercial_count),
      credit_unions = n_distinct(CU_NUMBER),
      .groups = "drop"
    ) |>
    mutate(year = year, average_size = safe_divide(amount, count), .before = geography)

  list(wi = wi_groups, geo = geo_groups)
})

cu_wi_rurality <- map_dfr(ncua_results, "wi") |>
  mutate(rurality = factor(rurality, levels = rurality_levels))

cu_geography <- map_dfr(ncua_results, "geo") |>
  mutate(geography = factor(geography, levels = c("Wisconsin", "United States")))

cu_wi_reference <- map_dfr(2017:2024, function(year) {
  wi_ids <- wi_cu_main_offices(year)$CU_NUMBER
  wi_fs <- load_ncua_fs(year) |> filter(CU_NUMBER %in% wi_ids)
  tibble(
    year,
    amount = sum(num0(wi_fs$ACCT_475A1)),
    count = sum(num0(wi_fs$ACCT_090A1))
  )
}) |>
  mutate(average_size = safe_divide(amount, count))

cu_wi_from_geo <- cu_geography |>
  filter(geography == "Wisconsin") |>
  arrange(year)

stopifnot(
  nrow(cu_wi_rurality) == 16L,
  nrow(cu_geography) == 16L,
  all(cu_wi_rurality$count > 0),
  all(cu_geography$count > 0),
  isTRUE(all.equal(
    cu_wi_reference$average_size,
    cu_wi_from_geo$average_size,
    check.attributes = FALSE
  )),
  all(
    cu_geography$amount[cu_geography$geography == "United States"] >=
      cu_geography$amount[cu_geography$geography == "Wisconsin"]
  )
)
```

```{r wi-cu-rurality-figure, fig.width=10, fig.height=5.5}
ggplot(
  cu_wi_rurality,
  aes(x = factor(year), y = average_size, fill = rurality)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    aes(label = short_dollar(average_size)),
    position = position_dodge(width = 0.8),
    vjust = -0.35,
    size = 2.8
  ) +
  scale_fill_manual(values = rurality_colors) +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Average Credit-Union Commercial Loan Size by Headquarters Rurality",
    subtitle = "Wisconsin-headquartered credit unions, ratio of group totals",
    x = "Year", y = "Average loan size", fill = NULL,
    caption = "Data: NCUA Call Reports; USDA Rural-Urban Continuum Codes"
  ) +
  charlie_theme() +
  theme(legend.position = "right")
```

# National comparisons

## National institutions and branches

```{r national-figure1, fig.width=10, fig.height=5.5}
national_fdic_long <- national_fdic |>
  transmute(year, Institutions = institutions, Branches = branches) |>
  pivot_longer(c(Institutions, Branches), names_to = "series", values_to = "value") |>
  mutate(series = factor(series, levels = c("Institutions", "Branches")))

ggplot(national_fdic_long, aes(year, value, color = series, shape = series)) +
  geom_line(linewidth = 1.05) +
  geom_point(size = 2.1) +
  scale_color_manual(values = c("Institutions" = "#BF4D28", "Branches" = "#2E75B6")) +
  scale_shape_manual(values = c("Institutions" = 16, "Branches" = 15)) +
  scale_x_continuous(breaks = seq(2000, 2024, 2)) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0.02, 0.08))) +
  labs(
    title = "U.S. Banking Institutions and Branches (2000–2024)",
    subtitle = "Institutions operating at least one branch in the 50 states or DC",
    x = "Year", y = "Count", color = NULL, shape = NULL,
    caption = "Data: FDIC Summary of Deposits"
  ) +
  charlie_theme() +
  theme(
    legend.position = "right",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

## National institutions by number of branches, 2024

```{r national-figure2-data}
national_sod_2024 <- load_fdic_sod(2024) |>
  filter(STALPBR %in% us_state_abbr)

national_institution_branches <- national_sod_2024 |>
  count(CERT, name = "branches")

institution_bucket_levels <- c("1", "2–5", "6–10", "11–20", "21–50", "50+")

national_figure2 <- national_institution_branches |>
  mutate(
    bucket = cut(
      branches,
      breaks = c(0, 2, 6, 11, 21, 51, Inf),
      labels = institution_bucket_levels,
      right = FALSE
    )
  ) |>
  count(bucket, name = "institutions") |>
  complete(
    bucket = factor(institution_bucket_levels, levels = institution_bucket_levels),
    fill = list(institutions = 0L)
  )

national_stats_text <- glue(
  "Mean: {round(mean(national_institution_branches$branches), 1)}  |  ",
  "Median: {round(median(national_institution_branches$branches), 0)}  |  ",
  "Total: {comma(nrow(national_institution_branches))}"
)

stopifnot(
  sum(national_figure2$institutions) == n_distinct(national_sod_2024$CERT),
  all(!is.na(national_figure2$bucket))
)
```

```{r national-figure2, fig.width=9.5, fig.height=5.5}
ggplot(national_figure2, aes(bucket, institutions)) +
  geom_col(fill = "#2E75B6", width = 0.7) +
  geom_text(aes(label = comma(institutions)), vjust = -0.35, size = 3.1) +
  annotate(
    "label", x = 6, y = max(national_figure2$institutions) * 0.92,
    label = national_stats_text, hjust = 1, size = 3, fill = "white"
  ) +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.12))) +
  labs(
    title = "U.S. Banking Institutions by Number of Branches (2024)",
    x = "Number of branches", y = "Number of institutions",
    caption = "Data: FDIC Summary of Deposits; geography is the 50 states plus DC"
  ) +
  charlie_theme()
```

## National lending by original loan size, 2023

```{r national-figure7}
national_figure7 <- cra_county |>
  filter(year == 2023L) |>
  build_figure7_table()

stopifnot(
  nrow(national_figure7) == 4L,
  abs(sum(national_figure7$loan_share[1:3]) - 1) < 1e-10,
  abs(sum(national_figure7$volume_share[1:3]) - 1) < 1e-10
)

kable(
  format_figure7(national_figure7),
  align = c("l", "r", "r", "r", "r", "r"),
  caption = "U.S. CRA small-business lending by original loan size (2023)"
)
```

## Lending frequency by geography and rurality

Each line is loans per 10,000 residents living in the same RUCC subgroup,
indexed to that line's own year-2000 value.

```{r cra-frequency-figure, fig.width=10.5, fig.height=11}
frequency_plot_data <- cra_normalized |>
  filter(measure == "Frequency")

ggplot(
  frequency_plot_data,
  aes(year, index, color = series, linetype = series)
) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey55") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(vars(bucket), ncol = 1) +
  scale_color_manual(values = comparison_colors, drop = FALSE) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotdash", "twodash"),
    drop = FALSE
  ) +
  scale_x_continuous(breaks = seq(2000, 2023, 2)) +
  scale_y_continuous(labels = label_number(accuracy = 1)) +
  labs(
    title = "CRA Small-Business Lending Frequency by Rurality",
    subtitle = "Loans per 10,000 subgroup residents; each series indexed to 2000 = 100",
    x = "Year", y = "Index (2000 = 100)", color = NULL, linetype = NULL,
    caption = paste(
      "Population: Census annual county resident estimates. Geography: 50 states + DC.",
      "RUCC 1–3 = urban/metro; RUCC 4–9 = rural/nonmetro."
    )
  ) +
  charlie_theme() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2))
```

## Nominal lending volume by geography and rurality

Each line is nominal loan dollars per 10,000 residents living in the same RUCC
subgroup, indexed to that line's own year-2000 value. Dollar amounts are not
inflation-adjusted, matching the report figure.

```{r cra-volume-figure, fig.width=10.5, fig.height=11}
volume_plot_data <- cra_normalized |>
  filter(measure == "Volume")

ggplot(
  volume_plot_data,
  aes(year, index, color = series, linetype = series)
) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey55") +
  geom_line(linewidth = 1) +
  geom_point(size = 1.6) +
  facet_wrap(vars(bucket), ncol = 1) +
  scale_color_manual(values = comparison_colors, drop = FALSE) +
  scale_linetype_manual(
    values = c("solid", "dashed", "dotdash", "twodash"),
    drop = FALSE
  ) +
  scale_x_continuous(breaks = seq(2000, 2023, 2)) +
  scale_y_continuous(labels = label_number(accuracy = 1)) +
  labs(
    title = "CRA Small-Business Lending Volume by Rurality",
    subtitle = "Nominal dollars per 10,000 subgroup residents; each series indexed to 2000 = 100",
    x = "Year", y = "Index (2000 = 100)", color = NULL, linetype = NULL,
    caption = paste(
      "Population: Census annual county resident estimates. Geography: 50 states + DC.",
      "Amounts are nominal. RUCC 1–3 = urban/metro; RUCC 4–9 = rural/nonmetro."
    )
  ) +
  charlie_theme() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.text = element_text(face = "bold")
  ) +
  guides(color = guide_legend(nrow = 2), linetype = guide_legend(nrow = 2))
```

```{r cra-index-endpoints}
kable(
  cra_normalized |>
    filter(year %in% c(2000L, 2023L)) |>
    select(year, series, measure, bucket, population, per_10k, index) |>
    mutate(
      population = comma(population, accuracy = 1),
      per_10k = number(per_10k, accuracy = 0.1),
      index = number(index, accuracy = 0.1)
    ),
  caption = "CRA subgroup normalization audit (endpoint years)"
)
```

## Wisconsin and U.S. credit-union commercial loan size

The U.S. group includes Wisconsin and contains credit unions headquartered in
the 50 states plus DC.

```{r cu-national-figure, fig.width=10, fig.height=5.5}
cu_geo_colors <- c("Wisconsin" = "#BF4D28", "United States" = "#2E75B6")

ggplot(
  cu_geography,
  aes(x = factor(year), y = average_size, fill = geography)
) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    aes(label = short_dollar(average_size)),
    position = position_dodge(width = 0.8),
    vjust = -0.35,
    size = 2.8
  ) +
  scale_fill_manual(values = cu_geo_colors) +
  scale_y_continuous(labels = label_dollar(scale_cut = cut_short_scale()),
                     expand = expansion(mult = c(0, 0.15))) +
  labs(
    title = "Average Credit-Union Commercial Loan Size",
    subtitle = "Wisconsin versus the United States, ratio of group totals",
    x = "Year", y = "Average loan size", fill = NULL,
    caption = "Data: NCUA Call Reports; U.S. geography is the 50 states plus DC"
  ) +
  charlie_theme() +
  theme(legend.position = "right")
```

```{r ncua-audit-table}
kable(
  cu_geography |>
    filter(year %in% c(2017L, 2024L)) |>
    mutate(
      amount = dollar(amount, accuracy = 1, scale_cut = cut_short_scale()),
      count = comma(count),
      average_size = dollar(average_size, accuracy = 1),
      credit_unions = comma(credit_unions)
    ),
  caption = "Credit-union commercial-loan audit (endpoint years)"
)
```

# Methods summary

- **Geography.** National outputs cover branch locations, CRA counties, or
  credit-union headquarters in the 50 states plus the District of Columbia.
  Wisconsin is included in the national benchmark.
- **Rurality.** RUCC 1–3 are labeled urban/metro and RUCC 4–9 rural/nonmetro.
  The lookup uses 2023 first, then 2013 and 2003 only for county FIPS absent
  from newer vintages. South Boston (`51780`) maps to Halifax County (`51083`).
- **Population.** CRA rates use annual Census county resident-population
  estimates aggregated within the same geography and RUCC subgroup as each
  numerator. Source series are final 2000–2010 and 2010–2020 intercensal
  estimates plus Vintage 2025 estimates for 2020–2023.
- **CRA amounts.** 2000–2004 and 2019–2023 source amounts are converted from
  thousands of dollars to dollars; 2005–2018 amounts are already dollars.
  Volume indexes remain nominal.
- **Credit-union averages.** Average commercial loan size is the ratio of total
  amount (`ACCT_475A1`) to total count (`ACCT_090A1`) within each plotted group,
  not an unweighted average of institution-level means.
