---
title: "National CRA Lending and Bank-Headquarters Trends"
author: "Codex"
date: 2026-08-24
format:
  html:
    embed-resources: true
    toc: true
execute:
  echo: false
  warning: false
  message: false
---

# Purpose

This workbook provides national benchmarks for two Wisconsin Community
Reinvestment Act (CRA) figures and for the decline in Wisconsin-headquartered
banks. The national geography is the 50 states plus the District of Columbia;
territories are excluded. All comparisons are descriptive.

The first two figures preserve the definitions and visual structure of current
banking Figures 8 and 9. They show CRA small-business lending frequency and
nominal lending volume per 10,000 denominator units for three original-loan-size
buckets, indexed so that 2000 equals 100. The third figure compares indexed
counts of Wisconsin-headquartered and nationally headquartered banks. The
fourth figure compares the mean number of branches per institution operating in
Wisconsin with the corresponding national mean.

```{r setup}
library(tidyverse)
library(scales)
library(knitr)

project_root <- normalizePath(file.path("..", ".."), winslash = "/")
knitr::opts_knit$set(root.dir = project_root)

source(file.path(
  project_root,
  "1_code", "1_2_visualize", "figs_charlie", "_charlie_helpers.R"
))

safe_divide <- function(num, den) {
  ifelse(is.na(den) | den == 0, NA_real_, num / den)
}

us_state_fips <- c(
  "01", "02", "04", "05", "06", "08", "09", "10", "11", "12", "13",
  "15", "16", "17", "18", "19", "20", "21", "22", "23", "24", "25",
  "26", "27", "28", "29", "30", "31", "32", "33", "34", "35", "36",
  "37", "38", "39", "40", "41", "42", "44", "45", "46", "47", "48",
  "49", "50", "51", "53", "54", "55", "56"
)

us_state_abbr <- sort(c(state.abb, "DC"))

stopifnot(
  length(us_state_fips) == 51L,
  n_distinct(us_state_fips) == 51L,
  "55" %in% us_state_fips,
  length(us_state_abbr) == 51L,
  n_distinct(us_state_abbr) == 51L,
  "WI" %in% us_state_abbr
)

required_file <- function(path) {
  if (!file.exists(path)) {
    stop("Missing required input file: ", path, call. = FALSE)
  }
  path
}

sum_numeric <- function(x) {
  sum(as.numeric(replace_na(x, 0)), na.rm = TRUE)
}
```

# Data construction and audits

The CRA source changes format twice during the analysis window. The loader below
handles each source era explicitly and applies the same unit conversions used by
the current Wisconsin figures. Amounts in 2000–2004 and 2019–2023 are reported
in thousands of dollars and are multiplied by 1,000. Amounts in 2005–2018 are
already denominated in dollars.

```{r cra-loader}
load_cra_series <- function(state_fips) {
  if (length(state_fips) == 0L || !all(state_fips %in% us_state_fips)) {
    stop("state_fips must contain eligible two-digit state/DC FIPS codes.")
  }

  records <- vector("list", length = 24L)

  for (yr in 2000:2023) {
    if (yr <= 2004) {
      path <- required_file(file.path(
        charlie_cra_dir, "aggr", sprintf("tract_%d.txt", yr)
      ))

      scoped <- readr::read_delim(
        path,
        delim = "|",
        show_col_types = FALSE,
        progress = FALSE
      ) |>
        mutate(
          state_fips = stringr::str_pad(as.character(state), 2, pad = "0")
        ) |>
        filter(report_level == 200, state_fips %in% !!state_fips)

      records[[yr - 1999L]] <- tibble(
        year = yr,
        loans_u100k_num = sum_numeric(scoped$num_100k),
        loans_100_250_num = sum_numeric(scoped$num_250k),
        loans_250_1m_num = sum_numeric(scoped$num_1mil),
        loans_u100k_amt = sum_numeric(scoped$vol_100k) * 1000,
        loans_100_250_amt = sum_numeric(scoped$vol_250k) * 1000,
        loans_250_1m_amt = sum_numeric(scoped$vol_1mil) * 1000
      )
    } else if (yr <= 2018) {
      path <- required_file(file.path(
        charlie_cra_dir, "cra_old", sprintf("cra_%d.csv", yr)
      ))

      scoped <- readr::read_csv(
        path,
        show_col_types = FALSE,
        progress = FALSE
      ) |>
        mutate(
          county_fips = stringr::str_pad(as.character(fips), 5, pad = "0"),
          state_fips = stringr::str_sub(county_fips, 1, 2)
        ) |>
        filter(state_fips %in% !!state_fips)

      records[[yr - 1999L]] <- tibble(
        year = yr,
        loans_u100k_num = sum_numeric(scoped$loan_count_100k),
        loans_100_250_num = sum_numeric(scoped$loan_count_250k),
        loans_250_1m_num = sum_numeric(scoped$loan_count_1M),
        loans_u100k_amt = sum_numeric(scoped$loan_vol_100k),
        loans_100_250_amt = sum_numeric(scoped$loan_vol_250k),
        loans_250_1m_amt = sum_numeric(scoped$loan_vol_1M)
      )
    } else {
      path <- required_file(file.path(
        charlie_cra_dir,
        sprintf("%dexp_aggr", yr - 2000L),
        sprintf("cra%d_Aggr_A11.dat", yr)
      ))

      scoped <- readr::read_fwf(
        path,
        col_positions = readr::fwf_widths(cra_fixed_widths, cra_fixed_names),
        show_col_types = FALSE,
        progress = FALSE
      ) |>
        mutate(
          state_fips = stringr::str_pad(as.character(State_FIPS), 2, pad = "0")
        ) |>
        filter(Report_Level == "200", state_fips %in% !!state_fips)

      records[[yr - 1999L]] <- tibble(
        year = yr,
        loans_u100k_num = sum_numeric(scoped$Loans_U100k_Num),
        loans_100_250_num = sum_numeric(scoped$Loans_100_250_Num),
        loans_250_1m_num = sum_numeric(scoped$Loans_250_1M_Num),
        loans_u100k_amt = sum_numeric(scoped$Loans_U100k_Amt) * 1000,
        loans_100_250_amt = sum_numeric(scoped$Loans_100_250_Amt) * 1000,
        loans_250_1m_amt = sum_numeric(scoped$Loans_250_1M_Amt) * 1000
      )
    }
  }

  bind_rows(records) |>
    arrange(year)
}

cra_national_raw <- load_cra_series(us_state_fips)
cra_wi_generic <- load_cra_series("55")

cra_value_cols <- setdiff(names(cra_national_raw), "year")

stopifnot(
  nrow(cra_national_raw) == 24L,
  identical(cra_national_raw$year, 2000:2023),
  !anyDuplicated(cra_national_raw$year),
  all(is.finite(as.matrix(cra_national_raw[cra_value_cols]))),
  all(as.matrix(cra_national_raw[cra_value_cols]) >= 0),
  all(
    as.matrix(cra_national_raw[cra_value_cols]) >=
      as.matrix(cra_wi_generic[cra_value_cols])
  )
)
```

The national denominator uses the same proxy as the current Wisconsin figures:
`Force / (Participation / 100)`. Geography membership is tested before any FIPS
padding or truncation. This excludes malformed extra rows `037` and `51000`
without turning them into duplicates of valid rows `37` and `51`.

```{r denominator-audit}
participation_raw <- readr::read_csv(
  required_file(file.path("0_inputs", "CORI", "fips_participation.csv")),
  col_types = readr::cols(
    FIPS = readr::col_character(),
    .default = readr::col_guess()
  ),
  progress = FALSE
) |>
  mutate(
    year = as.integer(year),
    Participation = readr::parse_number(as.character(Participation)),
    Force = readr::parse_number(as.character(Force))
  ) |>
  filter(year %in% 2000:2023)

eligible_population_rows <- participation_raw |>
  filter(FIPS %in% us_state_fips) |>
  mutate(population_proxy = safe_divide(Force, Participation / 100))

duplicate_population_keys <- eligible_population_rows |>
  count(year, FIPS) |>
  filter(n != 1L)

population_audit <- participation_raw |>
  group_by(year) |>
  summarise(
    eligible_geographies = n_distinct(FIPS[FIPS %in% us_state_fips]),
    eligible_rows = sum(FIPS %in% us_state_fips),
    excluded_rows = sum(!(FIPS %in% us_state_fips)),
    .groups = "drop"
  )

national_population <- eligible_population_rows |>
  group_by(year) |>
  summarise(
    population = sum(population_proxy),
    .groups = "drop"
  )

stopifnot(
  nrow(duplicate_population_keys) == 0L,
  nrow(population_audit) == 24L,
  all(population_audit$eligible_geographies == 51L),
  all(population_audit$eligible_rows == 51L),
  all(c("037", "51000") %in% participation_raw$FIPS),
  !any(c("037", "51000") %in% eligible_population_rows$FIPS),
  nrow(national_population) == 24L,
  all(national_population$population > 0)
)

knitr::kable(
  population_audit |>
    filter(year %in% c(2000L, 2023L)),
  caption = "National denominator geography audit (endpoint years)"
)
```

The generic Wisconsin CRA results are checked against the exact loaders used by
the current Figures 8 and 9. This ensures the national generalization has not
changed the existing bucket definitions or unit conversions.

```{r cra-validation}
wi_frequency_reference <- load_cra_wi_frequency_series()
wi_volume_reference <- load_cra_wi_volume_series()

frequency_matches <- isTRUE(all.equal(
  cra_wi_generic |>
    select(year, loans_u100k_num, loans_100_250_num, loans_250_1m_num),
  wi_frequency_reference,
  check.attributes = FALSE
))

volume_matches <- isTRUE(all.equal(
  cra_wi_generic |>
    select(year, loans_u100k_amt, loans_100_250_amt, loans_250_1m_amt),
  wi_volume_reference,
  check.attributes = FALSE
))

stopifnot(frequency_matches, volume_matches)

validation_summary <- tibble(
  check = c(
    "Generic Wisconsin frequency equals current Figure 8 loader",
    "Generic Wisconsin volume equals current Figure 9 loader",
    "National raw values are at least Wisconsin raw values",
    "National denominator has 51 eligible geographies each year"
  ),
  passed = c(
    frequency_matches,
    volume_matches,
    TRUE,
    all(population_audit$eligible_rows == 51L)
  )
)

knitr::kable(validation_summary, caption = "Core construction checks")
```

```{r build-cra-indexes}
cra_national <- cra_national_raw |>
  left_join(national_population, by = "year") |>
  arrange(year) |>
  mutate(
    loans_u100k_num_per_10k = loans_u100k_num / population * 10000,
    loans_100_250_num_per_10k = loans_100_250_num / population * 10000,
    loans_250_1m_num_per_10k = loans_250_1m_num / population * 10000,
    loans_u100k_amt_per_10k = loans_u100k_amt / population * 10000,
    loans_100_250_amt_per_10k = loans_100_250_amt / population * 10000,
    loans_250_1m_amt_per_10k = loans_250_1m_amt / population * 10000,
    loans_u100k_num_idx = loans_u100k_num_per_10k /
      loans_u100k_num_per_10k[year == 2000] * 100,
    loans_100_250_num_idx = loans_100_250_num_per_10k /
      loans_100_250_num_per_10k[year == 2000] * 100,
    loans_250_1m_num_idx = loans_250_1m_num_per_10k /
      loans_250_1m_num_per_10k[year == 2000] * 100,
    loans_u100k_amt_idx = loans_u100k_amt_per_10k /
      loans_u100k_amt_per_10k[year == 2000] * 100,
    loans_100_250_amt_idx = loans_100_250_amt_per_10k /
      loans_100_250_amt_per_10k[year == 2000] * 100,
    loans_250_1m_amt_idx = loans_250_1m_amt_per_10k /
      loans_250_1m_amt_per_10k[year == 2000] * 100
  )

cra_index_cols <- names(cra_national)[stringr::str_detect(
  names(cra_national), "_idx$"
)]

stopifnot(
  !anyNA(cra_national[cra_index_cols]),
  all(abs(as.numeric(cra_national[cra_national$year == 2000, cra_index_cols]) - 100) < 1e-10)
)
```

# National counterpart to Figure 8

```{r national-figure-8, fig.width=10, fig.height=5}
national_fig08_plot <- ggplot(cra_national, aes(x = year)) +
  geom_line(aes(y = loans_u100k_num_idx, color = "Under $100K"), linewidth = 1.1) +
  geom_point(aes(y = loans_u100k_num_idx, color = "Under $100K"), size = 2) +
  geom_line(aes(y = loans_100_250_num_idx, color = "$100K-$250K"), linewidth = 1.1) +
  geom_point(
    aes(y = loans_100_250_num_idx, color = "$100K-$250K"),
    shape = 15,
    size = 2
  ) +
  geom_line(aes(y = loans_250_1m_num_idx, color = "$250K-$1M"), linewidth = 1.1) +
  geom_point(
    aes(y = loans_250_1m_num_idx, color = "$250K-$1M"),
    shape = 17,
    size = 2
  ) +
  geom_hline(yintercept = 100, color = "gray50", linetype = "dashed") +
  scale_color_manual(
    values = c(
      "Under $100K" = "#2E75B6",
      "$100K-$250K" = "#BF4D28",
      "$250K-$1M" = "#4CAF50"
    ),
    breaks = c("Under $100K", "$100K-$250K", "$250K-$1M")
  ) +
  labs(
    title = "National Small Business Lending Frequency per 10K Residents",
    subtitle = "Growth Index (2000 = 100)",
    x = "Year",
    y = "Growth Index (2000 = 100)",
    color = NULL,
    caption = "Geography: 50 states and District of Columbia. Data: CRA aggregate files."
  ) +
  charlie_theme() +
  theme(legend.position = "top")

national_fig08_plot
```

The table below separates changes in national loan counts from changes in loans
per 10,000 denominator units. Percentage change is calculated as
`(2023 value - 2000 value) / 2000 value`, so positive values indicate growth
and negative values indicate decline. Figure 8 corresponds to the `Per-10K
change` column, not the raw `Count change` column.

```{r national-frequency-endpoints}
national_frequency_endpoints <- cra_national |>
  select(
    year,
    loans_u100k_num,
    loans_100_250_num,
    loans_250_1m_num,
    loans_u100k_num_per_10k,
    loans_100_250_num_per_10k,
    loans_250_1m_num_per_10k
  ) |>
  pivot_longer(
    cols = -year,
    names_to = c("bucket_code", ".value"),
    names_pattern = "(loans_u100k|loans_100_250|loans_250_1m)_(num_per_10k|num)$"
  ) |>
  mutate(
    bucket = recode(
      bucket_code,
      loans_u100k = "Under $100K",
      loans_100_250 = "$100K-$250K",
      loans_250_1m = "$250K-$1M"
    )
  ) |>
  filter(year %in% c(2000L, 2023L)) |>
  select(year, bucket, total_loans = num, loans_per_10k = num_per_10k) |>
  pivot_wider(
    names_from = year,
    values_from = c(total_loans, loans_per_10k),
    names_glue = "{.value}_{year}"
  ) |>
  mutate(
    total_loans_change = total_loans_2023 - total_loans_2000,
    total_loans_change_pct = safe_divide(
      total_loans_change,
      total_loans_2000
    ),
    loans_per_10k_change = loans_per_10k_2023 - loans_per_10k_2000,
    loans_per_10k_change_pct = safe_divide(
      loans_per_10k_change,
      loans_per_10k_2000
    )
  ) |>
  arrange(match(bucket, c("Under $100K", "$100K-$250K", "$250K-$1M")))

frequency_index_2023 <- cra_national |>
  filter(year == 2023L) |>
  transmute(
    `Under $100K` = loans_u100k_num_idx,
    `$100K-$250K` = loans_100_250_num_idx,
    `$250K-$1M` = loans_250_1m_num_idx
  ) |>
  unlist(use.names = FALSE)

stopifnot(all(abs(
  100 * (1 + national_frequency_endpoints$loans_per_10k_change_pct) -
    frequency_index_2023
) < 1e-10))

knitr::kable(
  national_frequency_endpoints |>
    transmute(
      bucket,
      `2000 loans` = scales::comma(total_loans_2000, accuracy = 1),
      `2023 loans` = scales::comma(total_loans_2023, accuracy = 1),
      `Count change` = scales::percent(
        total_loans_change_pct,
        accuracy = 0.1
      ),
      `2000 per 10K` = scales::number(loans_per_10k_2000, accuracy = 0.1),
      `2023 per 10K` = scales::number(loans_per_10k_2023, accuracy = 0.1),
      `Per-10K change` = scales::percent(
        loans_per_10k_change_pct,
        accuracy = 0.1
      )
    ),
  caption = "National lending-frequency endpoints by loan-size bucket"
)
```

# National counterpart to Figure 9

```{r national-figure-9, fig.width=10, fig.height=5}
national_fig09_plot <- ggplot(cra_national, aes(x = year)) +
  geom_line(aes(y = loans_u100k_amt_idx, color = "Under $100K"), linewidth = 1.1) +
  geom_point(aes(y = loans_u100k_amt_idx, color = "Under $100K"), size = 2) +
  geom_line(aes(y = loans_100_250_amt_idx, color = "$100K-$250K"), linewidth = 1.1) +
  geom_point(
    aes(y = loans_100_250_amt_idx, color = "$100K-$250K"),
    shape = 15,
    size = 2
  ) +
  geom_line(aes(y = loans_250_1m_amt_idx, color = "$250K-$1M"), linewidth = 1.1) +
  geom_point(
    aes(y = loans_250_1m_amt_idx, color = "$250K-$1M"),
    shape = 17,
    size = 2
  ) +
  geom_hline(yintercept = 100, color = "gray50", linetype = "dashed") +
  scale_color_manual(
    values = c(
      "Under $100K" = "#2E75B6",
      "$100K-$250K" = "#BF4D28",
      "$250K-$1M" = "#4CAF50"
    ),
    breaks = c("Under $100K", "$100K-$250K", "$250K-$1M")
  ) +
  labs(
    title = "National Small Business Lending Volume per 10K Residents",
    subtitle = "Growth Index (2000 = 100)",
    x = "Year",
    y = "Growth Index (2000 = 100)",
    color = NULL,
    caption = paste(
      "Nominal dollars; not adjusted for inflation.",
      "Geography: 50 states and District of Columbia. Data: CRA aggregate files."
    )
  ) +
  charlie_theme() +
  theme(legend.position = "top")

national_fig09_plot
```

The table below separates changes in nominal national totals from changes in
nominal dollars per 10,000 denominator units. Percentage change is calculated as
`(2023 value - 2000 value) / 2000 value`, so positive values indicate growth
and negative values indicate decline. Figure 9 corresponds to the `Per-10K
change` column, not the raw `Total change` column.

```{r national-volume-endpoints}
national_volume_endpoints <- cra_national |>
  select(
    year,
    loans_u100k_amt,
    loans_100_250_amt,
    loans_250_1m_amt,
    loans_u100k_amt_per_10k,
    loans_100_250_amt_per_10k,
    loans_250_1m_amt_per_10k
  ) |>
  pivot_longer(
    cols = -year,
    names_to = c("bucket_code", ".value"),
    names_pattern = "(loans_u100k|loans_100_250|loans_250_1m)_(amt_per_10k|amt)$"
  ) |>
  mutate(
    bucket = recode(
      bucket_code,
      loans_u100k = "Under $100K",
      loans_100_250 = "$100K-$250K",
      loans_250_1m = "$250K-$1M"
    )
  ) |>
  filter(year %in% c(2000L, 2023L)) |>
  select(year, bucket, nominal_total = amt, nominal_per_10k = amt_per_10k) |>
  pivot_wider(
    names_from = year,
    values_from = c(nominal_total, nominal_per_10k),
    names_glue = "{.value}_{year}"
  ) |>
  mutate(
    nominal_total_change = nominal_total_2023 - nominal_total_2000,
    nominal_total_change_pct = safe_divide(
      nominal_total_change,
      nominal_total_2000
    ),
    nominal_per_10k_change = nominal_per_10k_2023 - nominal_per_10k_2000,
    nominal_per_10k_change_pct = safe_divide(
      nominal_per_10k_change,
      nominal_per_10k_2000
    )
  ) |>
  arrange(match(bucket, c("Under $100K", "$100K-$250K", "$250K-$1M")))

volume_index_2023 <- cra_national |>
  filter(year == 2023L) |>
  transmute(
    `Under $100K` = loans_u100k_amt_idx,
    `$100K-$250K` = loans_100_250_amt_idx,
    `$250K-$1M` = loans_250_1m_amt_idx
  ) |>
  unlist(use.names = FALSE)

stopifnot(all(abs(
  100 * (1 + national_volume_endpoints$nominal_per_10k_change_pct) -
    volume_index_2023
) < 1e-10))

knitr::kable(
  national_volume_endpoints |>
    transmute(
      bucket,
      `2000 total` = scales::dollar(nominal_total_2000, accuracy = 1),
      `2023 total` = scales::dollar(nominal_total_2023, accuracy = 1),
      `Total change` = scales::percent(nominal_total_change_pct, accuracy = 0.1),
      `2000 per 10K` = scales::dollar(nominal_per_10k_2000, accuracy = 1),
      `2023 per 10K` = scales::dollar(nominal_per_10k_2023, accuracy = 1),
      `Per-10K change` = scales::percent(
        nominal_per_10k_change_pct,
        accuracy = 0.1
      )
    ),
  caption = "National nominal lending-volume endpoints by loan-size bucket"
)
```

# Wisconsin and national bank-headquarters trends

A bank is counted once by its FDIC certificate number when its Summary of
Deposits main-office record has `BRNUM == 0`. Wisconsin counts retain main
offices with `STALP == "WI"`; national counts retain main offices in the 50
states plus District of Columbia.

```{r headquarters-data}
headquarters_by_year <- purrr::map_dfr(2000:2024, function(target_year) {
  sod <- load_fdic_sod(target_year)

  main_offices <- sod |>
    filter(
      as.character(BRNUM) == "0",
      !is.na(CERT)
    )

  tibble(
    year = target_year,
    geography = c("Wisconsin", "United States"),
    hq_banks = c(
      main_offices |>
        filter(STALP == "WI") |>
        summarise(value = n_distinct(CERT)) |>
        pull(value),
      main_offices |>
        filter(STALP %in% us_state_abbr) |>
        summarise(value = n_distinct(CERT)) |>
        pull(value)
    )
  )
}) |>
  group_by(geography) |>
  arrange(year, .by_group = TRUE) |>
  mutate(
    hq_banks_idx = hq_banks / hq_banks[year == 2000] * 100
  ) |>
  ungroup()

wi_endpoint_reference <- purrr::map_dfr(c(2000L, 2024L), function(target_year) {
  tibble(
    year = target_year,
    hq_banks = load_fdic_sod(target_year) |>
      filter(
        as.character(BRNUM) == "0",
        STALP == "WI"
      ) |>
      distinct(CERT) |>
      nrow()
  )
})

stopifnot(
  nrow(headquarters_by_year) == 50L,
  n_distinct(headquarters_by_year$year) == 25L,
  all(headquarters_by_year$hq_banks > 0),
  all(headquarters_by_year$hq_banks == as.integer(headquarters_by_year$hq_banks)),
  all(
    headquarters_by_year |>
      filter(year == 2000L) |>
      pull(hq_banks_idx) == 100
  ),
  all(
    headquarters_by_year |>
      select(year, geography, hq_banks) |>
      pivot_wider(names_from = geography, values_from = hq_banks) |>
      transmute(passed = `United States` >= Wisconsin) |>
      pull(passed)
  ),
  isTRUE(all.equal(
    headquarters_by_year |>
      filter(geography == "Wisconsin", year %in% c(2000L, 2024L)) |>
      select(year, hq_banks),
    wi_endpoint_reference,
    check.attributes = FALSE
  ))
)

headquarters_endpoints <- headquarters_by_year |>
  filter(year %in% c(2000L, 2024L)) |>
  select(year, geography, hq_banks) |>
  pivot_wider(names_from = year, values_from = hq_banks, names_prefix = "banks_") |>
  mutate(
    decline = banks_2000 - banks_2024,
    decline_pct = safe_divide(decline, banks_2000)
  )
```

```{r headquarters-figure, fig.width=10, fig.height=5}
headquarters_plot <- ggplot(
  headquarters_by_year,
  aes(x = year, y = hq_banks_idx, color = geography, shape = geography)
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, color = "gray50", linetype = "dashed") +
  scale_color_manual(
    values = c("Wisconsin" = "#C5050C", "United States" = "#2E75B6"),
    breaks = c("Wisconsin", "United States")
  ) +
  scale_shape_manual(
    values = c("Wisconsin" = 16, "United States" = 15),
    breaks = c("Wisconsin", "United States")
  ) +
  labs(
    title = "Bank Headquarters in Wisconsin and the United States",
    subtitle = "Growth Index (2000 = 100)",
    x = "Year",
    y = "Growth Index (2000 = 100)",
    color = NULL,
    shape = NULL,
    caption = paste(
      "Institutions are distinct FDIC certificate numbers on main-office records.",
      "National geography is the 50 states plus District of Columbia."
    )
  ) +
  charlie_theme() +
  theme(legend.position = "top")

headquarters_plot
```

```{r headquarters-endpoint-table}
knitr::kable(
  headquarters_endpoints |>
    transmute(
      Geography = geography,
      `Banks in 2000` = scales::comma(banks_2000),
      `Banks in 2024` = scales::comma(banks_2024),
      `Absolute decline` = scales::comma(decline),
      `Percent decline` = scales::percent(decline_pct, accuracy = 0.1)
    ),
  caption = "Headquartered-bank endpoint comparison"
)
```

# Wisconsin and national branches per institution

This figure extends the mean series in the existing repository Figure 2c. For
each year and geography, institutions are distinct FDIC certificate numbers
with at least one branch located in that geography. The mean is total branch
records divided by the number of those institutions. Wisconsin therefore
includes all institutions operating branches in Wisconsin, regardless of
headquarters location. The national series includes branch locations in the 50
states plus District of Columbia.

```{r branches-per-institution-data}
branches_per_institution <- purrr::map_dfr(2000:2024, function(target_year) {
  sod <- load_fdic_sod(target_year)

  wi_branches <- sod |>
    filter(STALPBR == "WI")

  national_branches <- sod |>
    filter(STALPBR %in% us_state_abbr)

  wi_branch_counts <- wi_branches |>
    count(CERT, name = "institution_branches")

  tibble(
    year = target_year,
    geography = c("Wisconsin", "United States"),
    branches = c(nrow(wi_branches), nrow(national_branches)),
    institutions = c(
      n_distinct(wi_branches$CERT),
      n_distinct(national_branches$CERT)
    ),
    existing_figure_mean = c(
      mean(wi_branch_counts$institution_branches, na.rm = TRUE),
      NA_real_
    )
  ) |>
    mutate(
      branches_per_institution = safe_divide(branches, institutions)
    )
}) |>
  arrange(geography, year)

wi_mean_comparison <- branches_per_institution |>
  filter(geography == "Wisconsin") |>
  select(year, branches_per_institution, existing_figure_mean)

stopifnot(
  nrow(branches_per_institution) == 50L,
  n_distinct(branches_per_institution$year) == 25L,
  all(branches_per_institution$branches > 0),
  all(branches_per_institution$institutions > 0),
  !anyNA(branches_per_institution$branches_per_institution),
  all(abs(
    branches_per_institution$branches_per_institution -
      branches_per_institution$branches /
        branches_per_institution$institutions
  ) < 1e-12),
  all(abs(
    wi_mean_comparison$branches_per_institution -
      wi_mean_comparison$existing_figure_mean
  ) < 1e-12)
)

branches_per_institution_endpoints <- branches_per_institution |>
  filter(year %in% c(2000L, 2024L)) |>
  select(year, geography, branches_per_institution) |>
  pivot_wider(
    names_from = year,
    values_from = branches_per_institution,
    names_prefix = "ratio_"
  ) |>
  mutate(
    absolute_change = ratio_2024 - ratio_2000,
    percent_change = safe_divide(absolute_change, ratio_2000)
  )
```

```{r branches-per-institution-figure, fig.width=10, fig.height=5}
branches_per_institution_plot <- ggplot(
  branches_per_institution,
  aes(
    x = year,
    y = branches_per_institution,
    color = geography,
    shape = geography
  )
) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  scale_color_manual(
    values = c("Wisconsin" = "#C5050C", "United States" = "#2E75B6"),
    breaks = c("Wisconsin", "United States")
  ) +
  scale_shape_manual(
    values = c("Wisconsin" = 16, "United States" = 15),
    breaks = c("Wisconsin", "United States")
  ) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 2)) +
  labs(
    title = "Mean Branches per Banking Institution (2000-2024)",
    subtitle = "Institutions with branches operating in each geography",
    x = "Year",
    y = "Mean Branches per Institution",
    color = NULL,
    shape = NULL,
    caption = paste(
      "Data: FDIC Summary of Deposits.",
      "National geography is the 50 states plus District of Columbia."
    )
  ) +
  charlie_theme() +
  theme(
    legend.position = "top",
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

branches_per_institution_plot
```

```{r branches-per-institution-endpoint-table}
knitr::kable(
  branches_per_institution_endpoints |>
    transmute(
      Geography = geography,
      `Branches per institution, 2000` = scales::number(
        ratio_2000,
        accuracy = 0.01
      ),
      `Branches per institution, 2024` = scales::number(
        ratio_2024,
        accuracy = 0.01
      ),
      `Absolute change` = scales::number(
        absolute_change,
        accuracy = 0.01
      ),
      `Percent change` = scales::percent(percent_change, accuracy = 0.1)
    ),
  caption = "Mean branches-per-institution endpoint comparison"
)
```

# Descriptive findings and methods notes

```{r findings-values, include=FALSE}
wi_hq_result <- headquarters_endpoints |>
  filter(geography == "Wisconsin")

us_hq_result <- headquarters_endpoints |>
  filter(geography == "United States")

frequency_2023 <- cra_national |>
  filter(year == 2023L) |>
  select(
    loans_u100k_num_idx,
    loans_100_250_num_idx,
    loans_250_1m_num_idx
  )

volume_2023 <- cra_national |>
  filter(year == 2023L) |>
  select(
    loans_u100k_amt_idx,
    loans_100_250_amt_idx,
    loans_250_1m_amt_idx
  )

wi_branch_ratio_result <- branches_per_institution_endpoints |>
  filter(geography == "Wisconsin")

us_branch_ratio_result <- branches_per_institution_endpoints |>
  filter(geography == "United States")
```

The number of Wisconsin-headquartered banks fell from
`r scales::comma(wi_hq_result$banks_2000)` in 2000 to
`r scales::comma(wi_hq_result$banks_2024)` in 2024, a decline of
`r scales::percent(wi_hq_result$decline_pct, accuracy = 0.1)`. Nationally, the
comparable count fell from `r scales::comma(us_hq_result$banks_2000)` to
`r scales::comma(us_hq_result$banks_2024)`, a decline of
`r scales::percent(us_hq_result$decline_pct, accuracy = 0.1)`. The common
direction and the indexed paths in the third figure show whether the Wisconsin
trend broadly mirrors the national consolidation pattern; the endpoint
percentages quantify the difference in magnitude.

Mean branches per institution changed from
`r number(wi_branch_ratio_result$ratio_2000, accuracy = 0.01)` to
`r number(wi_branch_ratio_result$ratio_2024, accuracy = 0.01)` in Wisconsin and
from `r number(us_branch_ratio_result$ratio_2000, accuracy = 0.01)` to
`r number(us_branch_ratio_result$ratio_2024, accuracy = 0.01)` nationally. These
ratios describe the average branch footprint among institutions operating in
each geography; they are not restricted to institutions headquartered there.

In 2023, the national frequency indexes for the under-$100K, $100K–$250K, and
$250K–$1M buckets were respectively
`r number(frequency_2023$loans_u100k_num_idx, accuracy = 0.1)`,
`r number(frequency_2023$loans_100_250_num_idx, accuracy = 0.1)`, and
`r number(frequency_2023$loans_250_1m_num_idx, accuracy = 0.1)`. The corresponding
nominal-volume indexes were
`r number(volume_2023$loans_u100k_amt_idx, accuracy = 0.1)`,
`r number(volume_2023$loans_100_250_amt_idx, accuracy = 0.1)`, and
`r number(volume_2023$loans_250_1m_amt_idx, accuracy = 0.1)`. Values below 100
indicate that the 2023 per-10,000 measure was below its 2000 level.

These measures use the same denominator proxy as the current Wisconsin figures:
annual labor-force level divided by the labor-force participation rate. The CRA
volume series is nominal and is not adjusted for inflation. Results should
therefore be read as definition-matched descriptive benchmarks, not as causal
estimates.
