---
title: "What's driving rural Business Growth in Wisconsin?"
author: "Inder Majumdar"
format: html
---

```{r config, echo= FALSE, warning= FALSE, message= FALSE}
library(tidyverse)
library(plotly)

# Core processed inputs for workbook figures
bfs <- read_rds(file.path("..", "..", "2_processed_data", "BFS_county.rds")) |>
  mutate(across(everything(), ~ replace_na(.x, 0)))

cbp <- read_rds(file.path("..", "..", "2_processed_data", "CBP_all.rds"))

nes <- read_rds(file.path("..", "..", "2_processed_data", "NES_all.rds"))

# State-level participation data used to build population denominators
participation <- readr::read_csv(
  file.path("..", "..", "0_inputs", "CORI", "fips_participation.csv"),
  show_col_types = FALSE
) |>
  mutate(
    FIPS = stringr::str_pad(as.character(FIPS), width = 2, pad = "0"),
    Participation = readr::parse_number(as.character(Participation)),
    Force = readr::parse_number(as.character(Force)),
    year = as.integer(year),
    population = dplyr::if_else(Participation > 0, Force / (Participation / 100), NA_real_)
  ) |>
  select(FIPS, year, population)

# RUCC lookup for county-level metro/rural assignment
rucc <- readxl::read_excel(
  file.path("..", "..", "0_inputs", "Ruralurbancontinuumcodes2023.xlsx")
) |>
  mutate(
    FIPS = stringr::str_pad(as.character(FIPS), width = 5, side = "left", pad = "0"),
    RUCC_2023 = as.integer(RUCC_2023)
  ) |>
  select(FIPS, RUCC_2023)

midwest_excl_wi <- c("19", "17", "18", "26", "27")
big3 <- c("06", "25", "36")
wi_fips <- "55"

label_nat_excl <- "National (excl. CA, MA, NY)"
label_midwest <- "Midwest (excl. WI)"
label_wi <- "Wisconsin"
label_big3 <- "Big 3 (CA, MA, NY)"

category_colors <- c(
  "Wisconsin" = "#c5050c",
  "Midwest (excl. WI)" = "grey60",
  "National (excl. CA, MA, NY)" = "#6baed6",
  "Big 3 (CA, MA, NY)" = "#9e9ac8"
)

# Shared grouping helper used by all lollipop chunks
apply_grouping <- function(df, state_col) {
  df |>
    mutate(
      state_fips = stringr::str_pad(as.character(.data[[state_col]]), width = 2, pad = "0"),
      category = case_when(
        state_fips == wi_fips ~ label_wi,
        state_fips %in% midwest_excl_wi ~ label_midwest,
        state_fips %in% big3 ~ label_big3,
        TRUE ~ label_nat_excl
      )
    )
}

# Shared average-state aggregation helper used by raw-value lollipop chunks
avg_state_lollipop <- function(df, value_col) {
  df |>
    ungroup() |>
    group_by(state_fips, category) |>
    summarise(state_total = sum(.data[[value_col]], na.rm = TRUE), .groups = "drop") |>
    group_by(category) |>
    summarise(avg_state_value = mean(state_total, na.rm = TRUE), .groups = "drop") |>
    mutate(category = forcats::fct_reorder(category, avg_state_value))
}

# Shared average-state aggregation helper used by per-million lollipop chunks
avg_state_lollipop_per_million <- function(df, value_col) {
  years_use <- df |>
    distinct(year) |>
    pull(year) |>
    as.integer()

  state_totals <- df |>
    ungroup() |>
    group_by(state_fips, category) |>
    summarise(state_total = sum(.data[[value_col]], na.rm = TRUE), .groups = "drop")

  pop_totals <- participation |>
    filter(year %in% years_use) |>
    group_by(FIPS) |>
    summarise(sum_population = sum(population, na.rm = TRUE), .groups = "drop")

  state_totals |>
    left_join(pop_totals, by = c("state_fips" = "FIPS")) |>
    mutate(per_million = state_total / (sum_population / 1000000)) |>
    group_by(category) |>
    summarise(avg_state_value = mean(per_million, na.rm = TRUE), .groups = "drop") |>
    mutate(category = forcats::fct_reorder(category, avg_state_value))
}

# Prepped object for chunk lollipop_bfs
bfs_grouped <- apply_grouping(bfs, "state_fips") |>
  mutate(
    year = as.integer(year),
    business_app = suppressWarnings(as.numeric(business_app))
  )

# Prepped object for chunks lollipop_cbp and lollipop_cbp_1mill
cbp_grouped <- cbp |>
  mutate(
    year = as.integer(year),
    ESTAB = suppressWarnings(as.numeric(ESTAB)),
    naics_2digit = stringr::str_pad(as.character(naics), width = 2, pad = "0")
  ) |>
  apply_grouping("state")

# Prepped object for chunks lollipop_nes and lollipop_nes_1mill
nes_grouped <- nes |>
  mutate(
    year = as.integer(year),
    state = stringr::str_pad(as.character(state), width = 2, pad = "0"),
    county = stringr::str_pad(as.character(county), width = 3, pad = "0"),
    county_fips = paste0(state, county),
    NESTAB = suppressWarnings(as.numeric(NESTAB)),
    naics_2digit = stringr::str_pad(as.character(naics), width = 2, pad = "0")
  ) |>
  left_join(rucc, by = c("county_fips" = "FIPS")) |>
  mutate(
    rurality = if_else(RUCC_2023 %in% c(1, 2, 3), "metro", "rural")
  ) |>
  apply_grouping("state")

```


## Motivation

When analyzing the Business Formation Statistics (BFS) data in bfs_explore.html, we found that rural Wisconsin leads with respect to business applications between 2005 and 2024.


```{r lollipop_bfs, echo= FALSE, warning= FALSE, message= FALSE}
plot_data <- bfs_grouped |>
  filter(rurality == "rural") |>
  avg_state_lollipop("business_app")

ggplot(plot_data, aes(x = category, y = avg_state_value, color = category)) +
  geom_segment(aes(xend = category, y = 0, yend = avg_state_value), linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = category_colors) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(
    title = "BFS Business Applications (Average State)",
    subtitle = "Rural counties",
    x = NULL,
    y = "Average state business applications"
  )

```


This exact same pattern was found in the County Business Patterns (CBP) establishments data from 2010 to 2023, which provides more granular details on sector-wise establishment statistics (note to self: Was there a reason we didn't use the same time period, 2005-2024, as in BFS?)


```{r lollipop_cbp, echo= FALSE, warning= FALSE, message= FALSE}
plot_data <- cbp_grouped |>
  filter(rurality == "rural", naics_2digit == "00") |>
  avg_state_lollipop("ESTAB")

ggplot(plot_data, aes(x = category, y = avg_state_value, color = category)) +
  geom_segment(aes(xend = category, y = 0, yend = avg_state_value), linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = category_colors) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(
    title = "CBP Establishments (Average State)",
    subtitle = "Rural counties",
    x = NULL,
    y = "Average state establishments"
  )

```

## To what degree does population play a role?

Below is the CBP lollipop but adjusted per 1 million rural population.

```{r lollipop_cbp_1mill, echo= FALSE, warning= FALSE, message= FALSE}
plot_data <- cbp_grouped |>
  filter(rurality == "rural", naics_2digit == "00") |>
  avg_state_lollipop_per_million("ESTAB")

ggplot(plot_data, aes(x = category, y = avg_state_value, color = category)) +
  geom_segment(aes(xend = category, y = 0, yend = avg_state_value), linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = category_colors) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(
    title = "CBP Establishments per 1 Million Population (Average State)",
    subtitle = "Rural counties",
    x = NULL,
    y = "Average state establishments per 1 million"
  )

```


## Does the pattern hold using nonemployer statistics?

Nonemployer statistics are shown below using nonemployer establishments (`NESTAB`) and the same rural-only comparison framework.

```{r lollipop_nes, echo= FALSE, warning= FALSE, message= FALSE}
plot_data <- nes_grouped |>
  filter(rurality == "rural", naics_2digit == "00") |>
  avg_state_lollipop("NESTAB")

ggplot(plot_data, aes(x = category, y = avg_state_value, color = category)) +
  geom_segment(aes(xend = category, y = 0, yend = avg_state_value), linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = category_colors) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(
    title = "NES Establishments (Average State)",
    subtitle = "Rural counties",
    x = NULL,
    y = "Average state nonemployer establishments"
  )

```

```{r lollipop_nes_1mill, echo= FALSE, warning= FALSE, message= FALSE}
plot_data <- nes_grouped |>
  filter(rurality == "rural", naics_2digit == "00") |>
  avg_state_lollipop_per_million("NESTAB")

ggplot(plot_data, aes(x = category, y = avg_state_value, color = category)) +
  geom_segment(aes(xend = category, y = 0, yend = avg_state_value), linewidth = 1) +
  geom_point(size = 3) +
  scale_color_manual(values = category_colors) +
  coord_flip() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none") +
  labs(
    title = "NES Establishments per 1 Million Population (Average State)",
    subtitle = "Rural counties",
    x = NULL,
    y = "Average state nonemployer establishments per 1 million"
  )

```

## Using the NES Statistics, Where does Wisconsin Rank?


```{r Wisco_nes_rank, echo= FALSE, warning= FALSE, message= FALSE}
# Build yearly state totals for rural NES establishments per 1M population
state_lookup <- readr::read_csv(
  file.path("..", "..", "0_inputs", "state_fips.csv"),
  show_col_types = FALSE
) |>
  transmute(
    state_fips = stringr::str_pad(as.character(FIPS_CODE), width = 2, pad = "0"),
    state_name = stringr::str_to_title(STATE_NAME)
  )

nes_state_year <- nes_grouped |>
  filter(
    rurality == "rural",
    naics_2digit == "00"
  ) |>
  group_by(state_fips, year) |>
  summarise(
    nestab_total = sum(NESTAB, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(participation, by = c("state_fips" = "FIPS", "year" = "year")) |>
  mutate(nestab_per_million = nestab_total / (population / 1000000)) |>
  left_join(state_lookup, by = "state_fips") |>
  filter(!is.na(nestab_per_million), state_fips != "11", !is.na(state_name))

top10_plus_wi_nes <- function(df, value_col, label_col = "state_name") {
  ranked <- df |>
    filter(!is.na(.data[[value_col]])) |>
    arrange(desc(.data[[value_col]]))

  top10 <- ranked |> slice_head(n = 10)
  wi_row <- ranked |> filter(.data[[label_col]] == "Wisconsin")

  bind_rows(top10, wi_row) |>
    distinct(.data[[label_col]], .keep_all = TRUE) |>
    arrange(.data[[value_col]])
}

rank_years <- nes_state_year |>
  filter(state_name == "Wisconsin", !is.na(nestab_per_million)) |>
  distinct(year) |>
  pull(year) |>
  sort()

wi_nes_rank_data <- purrr::map_dfr(rank_years, function(y) {
  state_year_single <- nes_state_year |> filter(year == y)

  wi_rank_all <- state_year_single |>
    arrange(desc(nestab_per_million)) |>
    mutate(rank_all_states = row_number()) |>
    filter(state_name == "Wisconsin") |>
    pull(rank_all_states)

  top10_plus_wi_nes(state_year_single, "nestab_per_million") |>
    mutate(
      year = y,
      wi_rank_all_states = wi_rank_all,
      state_name = factor(state_name, levels = unique(state_name))
    )
})

wi_nes_rank_data <- wi_nes_rank_data |>
  mutate(year = as.integer(year))

years <- sort(unique(wi_nes_rank_data$year))
latest_year <- max(years, na.rm = TRUE)

frame_list <- purrr::map(years, function(y) {
  df <- wi_nes_rank_data |>
    filter(year == y) |>
    arrange(nestab_per_million) |>
    mutate(state_name = factor(state_name, levels = state_name))
  wi_row <- df |> filter(state_name == "Wisconsin")
  wi_rank <- unique(df$wi_rank_all_states)[1]

  list(
    name = as.character(y),
    data = list(
      list(
        x = df$nestab_per_million,
        y = df$state_name,
        text = scales::label_comma()(df$nestab_per_million),
        textposition = "outside",
        marker = list(color = ifelse(df$state_name == "Wisconsin", "#c5050c", "grey60"))
      )
    ),
    layout = list(
      yaxis = list(categoryorder = "array", categoryarray = levels(df$state_name)),
      annotations = list(
        list(
          x = 1.02,
          y = wi_row$state_name,
          xref = "paper",
          yref = "y",
          text = paste0("WI rank: #", wi_rank),
          showarrow = FALSE,
          xanchor = "left",
          yanchor = "middle",
          font = list(color = "#c5050c", size = 12)
        )
      )
    )
  )
})

init_df <- wi_nes_rank_data |>
  filter(year == latest_year) |>
  arrange(nestab_per_million) |>
  mutate(state_name = factor(state_name, levels = state_name))
init_wi <- init_df |> filter(state_name == "Wisconsin")
init_wi_rank <- unique(init_df$wi_rank_all_states)[1]
init_levels <- levels(init_df$state_name)

p <- plot_ly(
  init_df,
  x = ~nestab_per_million,
  y = ~state_name,
  text = ~scales::label_comma()(nestab_per_million),
  textposition = "outside",
  type = "bar",
  orientation = "h",
  hovertemplate = paste0(
    "State: %{y}<br>",
    "NESTAB per 1M: %{x:,.2f}<extra></extra>"
  ),
  marker = list(color = ~if_else(state_name == "Wisconsin", "#c5050c", "grey60"))
) |>
  layout(
    title = list(text = "Wisconsin NES Establishments per 1M (Top 10 + WI)"),
    xaxis = list(title = "NESTAB per 1M residents", tickformat = ","),
    yaxis = list(title = "", categoryorder = "array", categoryarray = init_levels),
    margin = list(l = 120, r = 130, t = 60, b = 40),
    annotations = list(
      list(
        x = 1.02,
        y = init_wi$state_name,
        xref = "paper",
        yref = "y",
        text = paste0("WI rank: #", init_wi_rank),
        showarrow = FALSE,
        xanchor = "left",
        yanchor = "middle",
        font = list(color = "#c5050c", size = 12)
      )
    ),
    showlegend = FALSE
  )

pb <- plotly_build(p)
pb$x$frames <- frame_list
pb$x$layout$sliders <- list(
  list(
    active = max(which(years == latest_year)) - 1,
    currentvalue = list(prefix = "Year: "),
    steps = purrr::map(years, function(y) {
      list(
        method = "animate",
        args = list(
          list(as.character(y)),
          list(mode = "immediate", frame = list(duration = 0, redraw = TRUE), transition = list(duration = 0))
        ),
        label = as.character(y)
      )
    })
  )
)

pb
```

```{r county_check, echo= FALSE, warning= FALSE, message= FALSE}

# bfs |>
#   select(county_fips_full) |>
#   unique() |>
#   nrow()
# 
# nes |>
#   mutate(county_fips_full = paste(state,county)) |>
#   select(county_fips_full) |>
#   unique() |>
#   nrow()
# 
# cbp |>
#   mutate(county_fips_full = paste(state,county)) |>
#   select(county_fips_full) |>
#   unique() |>
#   nrow()





```
