---
title: "2026-05-17 Report Sound-Bite Statistics"
author: "Codex"
format:
  html:
    embed-resources: true
date: 2026-05-17
---

# Purpose

This workbook computes a short list of high-signal baseline statistics that can be inserted into the Wisconsin small business finance report as reader-friendly "sound bites." The goal is to front-load a small number of concrete scale statistics in the opening paragraphs of the banking, lending, credit union, and equity sections.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(readxl)
library(glue)
library(scales)

project_root <- normalizePath(file.path("..", ".."), winslash = "/")
knitr::opts_knit$set(root.dir = project_root)
source(file.path(project_root, "1_code", "1_2_visualize", "figs_charlie", "_charlie_helpers.R"))

safe_divide <- function(num, den) {
  ifelse(is.na(den) | den == 0, NA_real_, num / den)
}
```

# Banking Structure

```{r banking-structure}
bank_2000 <- load_fdic_sod(2000) |>
  filter(STALPBR == "WI")

bank_2024 <- load_fdic_sod(2024) |>
  filter(STALPBR == "WI")

banking_stats <- tibble(
  institutions_2000 = n_distinct(bank_2000$CERT),
  branches_2000 = nrow(bank_2000),
  institutions_2024 = n_distinct(bank_2024$CERT),
  branches_2024 = nrow(bank_2024),
  top5_branch_share_2024 = bank_2024 |>
    count(CERT, sort = TRUE, name = "branches") |>
    slice_head(n = 5) |>
    summarise(share = sum(branches) / nrow(bank_2024)) |>
    pull(share)
)

banking_stats
```

# Wisconsin-Headquartered Banks

This section identifies Wisconsin-headquartered banks using FDIC main-office
records (`BRNUM == 0` and `STALP == "WI"`). The corresponding branch count is
restricted to Wisconsin branch offices operated by those institutions
(`STALPBR == "WI"`).

```{r wi-headquartered-banks}
wi_hq_by_year <- map_dfr(c(2000L, 2024L), function(target_year) {
  sod <- load_fdic_sod(target_year)

  wi_hq_certs <- sod |>
    filter(
      as.character(BRNUM) == "0",
      STALP == "WI"
    ) |>
    distinct(CERT)

  tibble(
    year = target_year,
    hq_banks = nrow(wi_hq_certs),
    wi_branches_of_hq_banks = sod |>
      filter(
        STALPBR == "WI",
        CERT %in% wi_hq_certs$CERT
      ) |>
      nrow()
  )
})

wi_hq_stats <- wi_hq_by_year |>
  summarise(
    hq_banks_2000 = hq_banks[year == 2000][1],
    hq_banks_2024 = hq_banks[year == 2024][1],
    hq_banks_decline = hq_banks_2000 - hq_banks_2024,
    hq_banks_decline_share = safe_divide(
      hq_banks_decline,
      hq_banks_2000
    ),
    wi_branches_2000 = wi_branches_of_hq_banks[year == 2000][1],
    wi_branches_2024 = wi_branches_of_hq_banks[year == 2024][1],
    wi_branches_decline = wi_branches_2000 - wi_branches_2024,
    wi_branches_decline_share = safe_divide(
      wi_branches_decline,
      wi_branches_2000
    )
  )

wi_hq_soundbite <- glue(
  "Between 2000 and 2024, the number of banks headquartered in Wisconsin ",
  "fell from {comma(wi_hq_stats$hq_banks_2000)} to ",
  "{comma(wi_hq_stats$hq_banks_2024)}, a decline of ",
  "{percent(wi_hq_stats$hq_banks_decline_share, accuracy = 0.1)}. ",
  "Over the same period, the number of Wisconsin branches operated by these ",
  "banks fell from {comma(wi_hq_stats$wi_branches_2000)} to ",
  "{comma(wi_hq_stats$wi_branches_2024)}, a decline of ",
  "{percent(wi_hq_stats$wi_branches_decline_share, accuracy = 0.1)}."
)

wi_hq_by_year
wi_hq_stats
wi_hq_soundbite
```

# Small Business Lending Snapshot

```{r cra-figure7}
cra_2023 <- readr::read_fwf(
  file.path(charlie_cra_dir, "23exp_aggr", "cra2023_Aggr_A11.dat"),
  col_positions = readr::fwf_widths(cra_fixed_widths, cra_fixed_names),
  show_col_types = FALSE,
  progress = FALSE
) |>
  filter(State_FIPS == "55", Report_Level == "200") |>
  mutate(
    Loans_U100k_Num = as.numeric(replace_na(Loans_U100k_Num, "0")),
    Loans_U100k_Amt = as.numeric(replace_na(Loans_U100k_Amt, "0")) * 1000,
    Loans_100_250_Num = as.numeric(replace_na(Loans_100_250_Num, "0")),
    Loans_100_250_Amt = as.numeric(replace_na(Loans_100_250_Amt, "0")) * 1000,
    Loans_250_1M_Num = as.numeric(replace_na(Loans_250_1M_Num, "0")),
    Loans_250_1M_Amt = as.numeric(replace_na(Loans_250_1M_Amt, "0")) * 1000
  )

cra_stats <- tibble(
  total_loans_2023 = sum(
    cra_2023$Loans_U100k_Num +
      cra_2023$Loans_100_250_Num +
      cra_2023$Loans_250_1M_Num,
    na.rm = TRUE
  ),
  total_volume_2023 = sum(
    cra_2023$Loans_U100k_Amt +
      cra_2023$Loans_100_250_Amt +
      cra_2023$Loans_250_1M_Amt,
    na.rm = TRUE
  ),
  avg_loan_size_2023 = safe_divide(
    sum(
      cra_2023$Loans_U100k_Amt +
        cra_2023$Loans_100_250_Amt +
        cra_2023$Loans_250_1M_Amt,
      na.rm = TRUE
    ),
    sum(
      cra_2023$Loans_U100k_Num +
        cra_2023$Loans_100_250_Num +
        cra_2023$Loans_250_1M_Num,
      na.rm = TRUE
    )
  ),
  u100k_num_share = safe_divide(sum(cra_2023$Loans_U100k_Num, na.rm = TRUE),
                                sum(cra_2023$Loans_U100k_Num + cra_2023$Loans_100_250_Num + cra_2023$Loans_250_1M_Num, na.rm = TRUE)),
  u100k_amt_share = safe_divide(sum(cra_2023$Loans_U100k_Amt, na.rm = TRUE),
                                sum(cra_2023$Loans_U100k_Amt + cra_2023$Loans_100_250_Amt + cra_2023$Loans_250_1M_Amt, na.rm = TRUE)),
  hi_num_share = safe_divide(sum(cra_2023$Loans_250_1M_Num, na.rm = TRUE),
                             sum(cra_2023$Loans_U100k_Num + cra_2023$Loans_100_250_Num + cra_2023$Loans_250_1M_Num, na.rm = TRUE)),
  hi_amt_share = safe_divide(sum(cra_2023$Loans_250_1M_Amt, na.rm = TRUE),
                             sum(cra_2023$Loans_U100k_Amt + cra_2023$Loans_100_250_Amt + cra_2023$Loans_250_1M_Amt, na.rm = TRUE))
)

cra_stats
```

# Credit Union Snapshot

```{r credit-union-snapshot}
cu_branches_2024 <- load_ncua_branch(2024) |>
  filter(PhysicalAddressStateCode == "WI")

cu_hq_2023 <- load_ncua_branch(2023) |>
  filter(
    PhysicalAddressStateCode == "WI",
    MainOffice == "Yes"
  )

cu_stats <- tibble(
  institutions_2024 = n_distinct(cu_branches_2024$CU_NUMBER),
  branches_2024 = nrow(cu_branches_2024),
  hq_rows_2023 = nrow(cu_hq_2023),
  hq_counties_2023 = cu_hq_2023 |>
    distinct(PhysicalAddressCountyName) |>
    nrow(),
  no_hq_counties_2023 = 72 - (cu_hq_2023 |>
    distinct(PhysicalAddressCountyName) |>
    nrow())
)

cu_stats
```

# Credit Union Share of Branch Locations

This section compares Wisconsin bank and credit union branch records for 2024.
The denominator is the combined number of FDIC-reported bank branches and
NCUA-reported credit union branches located in Wisconsin.

```{r credit-union-branch-share}
branch_location_mix <- tibble(
  bank_branches_2024 = nrow(bank_2024),
  cu_branches_2024 = nrow(cu_branches_2024)
) |>
  mutate(
    combined_branches_2024 = bank_branches_2024 + cu_branches_2024,
    cu_branch_share_2024 = safe_divide(
      cu_branches_2024,
      combined_branches_2024
    )
  )

cu_branch_share_soundbite <- glue(
  "In 2024, credit unions operated ",
  "{comma(branch_location_mix$cu_branches_2024)} of Wisconsin's ",
  "{comma(branch_location_mix$combined_branches_2024)} combined bank and ",
  "credit union branch locations, accounting for ",
  "{percent(branch_location_mix$cu_branch_share_2024, accuracy = 0.1)} ",
  "of the total."
)

branch_location_mix
cu_branch_share_soundbite
```

# Credit Union Commercial Lending Growth Versus Bank Lending Decline

This section compares nominal annual lending volumes beginning in 2001, the
first year available in the staged NCUA archive. For credit unions, the 2001
measure is total member business loans granted year to date (`ACCT_475`) and
the 2024 measure is total member commercial loans granted or purchased year to
date (`ACCT_475A1`). Both totals are restricted to Wisconsin-headquartered
credit unions. The bank comparison uses CRA-reported originations of loans of
$1 million or less and ends in 2023, the latest staged CRA year. These series
are directionally informative but not directly equivalent because their
reporting definitions and end years differ.

```{r credit-union-commercial-lending-growth}
cu_2001_dir <- file.path(charlie_ncua_dir, "QCR200112")

cu_2001_institutions <- readr::read_csv(
  file.path(cu_2001_dir, "FOICU.txt"),
  show_col_types = FALSE,
  progress = FALSE,
  locale = readr::locale(encoding = "cp1252")
) |>
  filter(STATE == "WI") |>
  distinct(CU_NUMBER)

cu_commercial_lending <- bind_rows(
  readr::read_csv(
    file.path(cu_2001_dir, "fs220A.txt"),
    show_col_types = FALSE,
    progress = FALSE,
    locale = readr::locale(encoding = "cp1252"),
    guess_max = 50000
  ) |>
    filter(CU_NUMBER %in% cu_2001_institutions$CU_NUMBER) |>
    summarise(
      year = 2001L,
      annual_commercial_lending = sum(as.numeric(ACCT_475), na.rm = TRUE)
    ),
  load_ncua_fs(2024) |>
    filter(CU_NUMBER %in% wi_cu_main_offices(2024)$CU_NUMBER) |>
    summarise(
      year = 2024L,
      annual_commercial_lending = sum(
        as.numeric(ACCT_475A1),
        na.rm = TRUE
      )
    )
) |>
  arrange(year)

bank_small_business_lending <- load_cra_wi_volume_series() |>
  transmute(
    year,
    annual_small_business_lending =
      loans_u100k_amt + loans_100_250_amt + loans_250_1m_amt
  ) |>
  filter(year %in% c(2001L, 2023L))

cu_bank_lending_stats <- tibble(
  cu_volume_2001 = cu_commercial_lending |>
    filter(year == 2001L) |>
    pull(annual_commercial_lending),
  cu_volume_2024 = cu_commercial_lending |>
    filter(year == 2024L) |>
    pull(annual_commercial_lending),
  bank_volume_2001 = bank_small_business_lending |>
    filter(year == 2001L) |>
    pull(annual_small_business_lending),
  bank_volume_2023 = bank_small_business_lending |>
    filter(year == 2023L) |>
    pull(annual_small_business_lending)
) |>
  mutate(
    cu_volume_increase = cu_volume_2024 - cu_volume_2001,
    cu_volume_increase_share = safe_divide(
      cu_volume_increase,
      cu_volume_2001
    ),
    bank_volume_decline = bank_volume_2001 - bank_volume_2023,
    bank_volume_decline_share = safe_divide(
      bank_volume_decline,
      bank_volume_2001
    ),
    cu_increase_as_share_of_bank_decline = safe_divide(
      cu_volume_increase,
      bank_volume_decline
    )
  )

cu_bank_lending_soundbite <- glue(
  "Reported annual commercial lending by Wisconsin-headquartered credit ",
  "unions rose from {dollar(cu_bank_lending_stats$cu_volume_2001, ",
  "accuracy = 0.1, scale = 1e-6, suffix = ' million')} in 2001 to ",
  "{dollar(cu_bank_lending_stats$cu_volume_2024, accuracy = 0.01, ",
  "scale = 1e-9, suffix = ' billion')} in 2024, an increase of ",
  "{dollar(cu_bank_lending_stats$cu_volume_increase, accuracy = 0.01, ",
  "scale = 1e-9, suffix = ' billion')}, or ",
  "{percent(cu_bank_lending_stats$cu_volume_increase_share, accuracy = 0.1)}. ",
  "Over the closest available CRA comparison, Wisconsin bank originations ",
  "of loans of $1 million or less fell from ",
  "{dollar(cu_bank_lending_stats$bank_volume_2001, accuracy = 0.01, ",
  "scale = 1e-9, suffix = ' billion')} in 2001 to ",
  "{dollar(cu_bank_lending_stats$bank_volume_2023, accuracy = 0.01, ",
  "scale = 1e-9, suffix = ' billion')} in 2023, a decline of ",
  "{dollar(cu_bank_lending_stats$bank_volume_decline, accuracy = 0.01, ",
  "scale = 1e-9, suffix = ' billion')}, or ",
  "{percent(cu_bank_lending_stats$bank_volume_decline_share, accuracy = 0.1)}. ",
  "The credit union increase was equal to ",
  "{percent(cu_bank_lending_stats$cu_increase_as_share_of_bank_decline, ",
  "accuracy = 0.1)} of the bank decline in nominal dollar terms."
)

cu_commercial_lending
bank_small_business_lending
cu_bank_lending_stats
cu_bank_lending_soundbite
```

# Venture Capital Snapshot

```{r vc-snapshot}
year_min <- 2015
year_max <- 2024
wi_fips <- "55"

count_raw <- readxl::read_xlsx(file.path("0_inputs", "Pitchbook", "Pitchbook_dealcount.xlsx"))
vol_raw <- readxl::read_xlsx(file.path("0_inputs", "Pitchbook", "Pitchbook_dealvol.xlsx"))

participation <- readr::read_csv(
  file.path("0_inputs", "CORI", "fips_participation.csv"),
  show_col_types = FALSE
) |>
  mutate(
    FIPS = stringr::str_pad(as.character(FIPS), width = 2, pad = "0"),
    Participation = readr::parse_number(as.character(Participation)),
    Force = readr::parse_number(as.character(Force)),
    year = as.integer(year),
    population = safe_divide(Force, Participation / 100)
  ) |>
  select(FIPS, year, population)

state_fips <- readr::read_csv(
  file.path("0_inputs", "state_fips.csv"),
  show_col_types = FALSE
) |>
  mutate(
    FIPS = stringr::str_pad(as.character(FIPS_CODE), width = 2, pad = "0"),
    state_name_key = tolower(STATE_NAME)
  ) |>
  select(FIPS, state_name_key)

count_long <- count_raw |>
  pivot_longer(-State, names_to = "year", values_to = "dealcount") |>
  mutate(year = as.integer(year)) |>
  filter(year >= year_min, year <= year_max)

vol_long <- vol_raw |>
  pivot_longer(-State, names_to = "year", values_to = "capital_committed") |>
  mutate(year = as.integer(year)) |>
  filter(year >= year_min, year <= year_max)

state_year <- count_long |>
  left_join(vol_long, by = c("State", "year")) |>
  mutate(state_name = State, state_name_key = tolower(state_name)) |>
  left_join(state_fips, by = "state_name_key") |>
  left_join(participation, by = c("FIPS", "year")) |>
  filter(!is.na(FIPS), FIPS != "11")

state_period <- state_year |>
  group_by(state_name, FIPS) |>
  summarise(
    sum_dealcount = sum(dealcount, na.rm = TRUE),
    sum_capital = sum(capital_committed, na.rm = TRUE),
    sum_population = sum(population, na.rm = TRUE),
    committed_per_million = safe_divide(sum_capital, sum_population / 1000000),
    .groups = "drop"
  )

wi_period <- state_period |>
  filter(FIPS == wi_fips)

vc_stats <- tibble(
  total_vc_2015_2024 = wi_period$sum_capital,
  avg_annual_vc = wi_period$sum_capital / 10,
  share_us_total = wi_period$sum_capital / sum(state_period$sum_capital, na.rm = TRUE),
  committed_per_million = wi_period$committed_per_million,
  rank_contig45 = state_period |>
    filter(!state_name %in% c("California", "New York", "Massachusetts", "Delaware")) |>
    arrange(desc(committed_per_million)) |>
    mutate(rank = row_number()) |>
    filter(FIPS == wi_fips) |>
    pull(rank)
)

vc_stats
```

# Approved Stat List and Draft Insertions

```{r approved-list}
approved_stats <- tibble(
  section = c(
    "Banking structure opener",
    "Small business lending opener",
    "Credit union market snapshot",
    "Equity / venture capital opener"
  ),
  place_in_draft = c(
    "Insert as the second sentence of paragraph 19, immediately after `Wisconsin’s banking sector has undergone consolidation over the past two decades...`",
    "Insert as the second sentence of paragraph 38, immediately after `Since accessibility to banking services has changed considerably since 2000...`",
    "Insert after paragraph 74, following the sentence ending `...whether that activity aligns with areas where access to traditional banking services has declined.`",
    "Insert after paragraph 86, immediately after `We now turn our focus to equity financing.`"
  ),
  suggested_text = c(
    glue("As of 2024, Wisconsin had {comma(banking_stats$institutions_2024)} banking institutions operating {comma(banking_stats$branches_2024)} branches statewide, down from {comma(banking_stats$institutions_2000)} institutions and {comma(banking_stats$branches_2000)} branches in 2000."),
    glue("In 2023, Wisconsin lenders originated {comma(cra_stats$total_loans_2023)} CRA-reported small business loans totaling {dollar(cra_stats$total_volume_2023 / 1e9, accuracy = 0.1, suffix = 'B')}, with an average loan size of about {dollar(cra_stats$avg_loan_size_2023, accuracy = 1)}."),
    glue("As of 2024, Wisconsin had {comma(cu_stats$institutions_2024)} credit unions operating {comma(cu_stats$branches_2024)} branches statewide. Yet in 2023, credit union main offices were present in only {comma(cu_stats$hq_counties_2023)} of Wisconsin’s 72 counties, leaving {comma(cu_stats$no_hq_counties_2023)} counties without a local credit union headquarters."),
    glue("Between 2015 and 2024, Wisconsin firms raised about {dollar(vc_stats$total_vc_2015_2024, accuracy = 0.01, scale = 1000, suffix = ' billion')} in venture capital, an average of {dollar(vc_stats$avg_annual_vc, accuracy = 1, suffix = ' million')} per year, accounting for roughly {percent(vc_stats$share_us_total, accuracy = 0.01)} of U.S. venture funding over the period.")
  )
)

approved_stats
```

# Notes

- The banking and CRA calculations are built directly from the same local FDIC and CRA inputs already used in the figure scripts.
- Wisconsin-headquartered banks are identified from FDIC main-office records using `BRNUM == 0` and `STALP == "WI"`; the associated branch counts include only their branch offices located in Wisconsin.
- The bank-versus-credit-union branch share uses 2024 Wisconsin branch records from FDIC SOD and NCUA; it does not deduplicate locations shared by separate institutions.
- The credit union snapshot uses the same NCUA branch files as the CU figures, with headquarters proxied by `MainOffice == "Yes"`.
- The venture capital snapshot follows the same PitchBook plus participation-data logic already used in `1_code/workbooks/2026_02_08_pitchbook.qmd`.
