---
title: "What's driving rural Business Growth in Wisconsin?"
author: "Inder Majumdar"
format: html
---

```{r config, echo= FALSE, warning= FALSE, message= FALSE}
library(tidyverse)
library(plotly)

bfs <- read_rds(file.path("..", "..", "2_processed_data", "BFS_county.rds")) |>
  mutate(across(everything(), ~ replace_na(.x, 0)))

cbp <- read_rds(file.path("..", "..", "2_processed_data", "CBP_all.rds"))

participation <- readr::read_csv(
  file.path("..", "..", "0_inputs", "CORI", "fips_participation.csv"),
  show_col_types = FALSE
) |>
  mutate(
    FIPS = stringr::str_pad(as.character(FIPS), width = 2, pad = "0"),
    Participation = readr::parse_number(as.character(Participation)),
    Force = readr::parse_number(as.character(Force)),
    year = as.integer(year),
    population = dplyr::if_else(Participation > 0, Force / (Participation / 100), NA_real_)
  ) |>
  select(FIPS, year, population)

naics_lookup <- readr::read_csv(
  file.path("..", "..", "0_inputs", "naics_2digit_sectors_2007_2012_2017.csv"),
  show_col_types = FALSE
) |>
  mutate(
    year = as.integer(year),
    naics_2digit = stringr::str_pad(as.character(naics_2digit), width = 2, pad = "0")
  ) |>
  distinct(year, naics_2digit, description)

midwest_excl_wi <- c("19", "17", "18", "26", "27")
big3 <- c("06", "25", "36")
wi_fips <- "55"

label_nat <- "National"
label_nat_excl <- "National (excl. CA, MA, NY)"
label_midwest <- "Midwest (excl. WI)"
label_wi <- "Wisconsin"
label_big3 <- "Big 3 (CA, MA, NY)"

category_colors <- c(
  "Wisconsin" = "#c5050c",
  "Midwest (excl. WI)" = "grey60",
  "National (excl. CA, MA, NY)" = "#6baed6",
  "National" = "#08519c",
  "Big 3 (CA, MA, NY)" = "#9e9ac8"
)

nes <- read_rds(file.path("..", "..", "2_processed_data", "NES_all.rds"))

```


## Motivation

When analyzing the Business Formation Statistics (BFS) data in bfs_explore.html, we found that rural Wisconsin leads with respect to business applications between 2005 and 2024.


```{r lollipop_bfs, echo= FALSE, warning= FALSE, message= FALSE}



```


This exact same pattern was found in the County Business Patterns (CBP) establishments data from 2010 to 2023, which provides more granular details on sector-wise establishment statistics (note to self: Was there a reason we didn't use the same time period, 2005-2024, as in BFS?)


```{r lollipop_cbp, echo= FALSE, warning= FALSE, message= FALSE}



```

## To what degree does population play a role?

Below is the CBP lollipop but adjusted per 1 million rural population. Population statistics are collected

```{r lollipop_cbp_1mill, echo= FALSE, warning= FALSE, message= FALSE}


```


## Does the pattern hold using nonemployer statistics?

Nonemployer statistics are found 
